### 2018-05-25 Mouse_Microbiome.rmd
Work from 5.25, playing with mortality

---
title: "Mouse_Microbiome"
author: "Tim"
date: "5/25/2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE)
#set working directory
setwd("/Users/sgamoney/Google Drive/mouse_microbiome/Mouse_Microbiome")

#load necessary packages -- "library" or "require"
require(tidyverse)
require(data.table)

#load data
abun_raw <- fread("2016_10_13_run2_otus.csv")
abun <- abun_raw
abun[345,4719] <- 0
abun <- abun %>%
  select(-Run, -Day, -Reads) %>%
  rename(Cage = `Cage#`) %>%
  gather(otuid, abun, -SampleID, -Cage, - Abx, -Dose, -Hour, -Cohoused, -Pair) %>%
  group_by(SampleID) %>%
  mutate(abun = abun/sum(abun))
```

Okay, now we'll learn to make *plots*!

```{r demo plot}
#demo plots
tempz <- abun %>%
  group_by(otuid, Hour) %>%
  summarise(abun = sum(abun)) %>%
  group_by(otuid) %>%
  mutate(total = sum(abun)) %>%
  filter(total > 7)

ggplot(tempz, aes(x = Hour, y = abun, color = otuid)) +
  geom_point() +
  stat_smooth() +
  facet_wrap(~otuid, scale = 'free_y')
```

```{r initial mouse mortality}

#set working directory
setwd("/Users/sgamoney/Google Drive/mouse_microbiome/Mouse_Microbiome")

#load necessary packages -- "library" or "require"
require(tidyverse)
require(data.table)

#load and set data
raw_abun <- fread("2016_10_13_run2_otus.csv")
new_abun <- raw_abun
new_abun[345,4719] <- 0
new_abun <- new_abun %>%
  select(-Run, -Day, -Reads) %>%
  rename(Cage = `Cage#`) %>%
  gather(otuid, abun, -SampleID, -Cage, -Abx, -Dose, -Hour, -Cohoused, -Pair) %>%
  group_by(SampleID) %>%
  mutate(abun = abun/sum(abun)) %>%
  mutate(mouseid = paste(Cage, Abx, sep = "_"))

#Filter time before 120 Hours and assign one of three mortality codes 
new_abun <- new_abun %>%
  filter(Hour < 121) %>%
  group_by(mouseid, otuid) %>%
  summarise(
    mstatus = ifelse(sum(abun) == 0, "absent", 
                     ifelse(abun[Hour == 120] > 0, "persistent", "leaves"))) 
```

### 2018-06-12 Initial Community Composition.Rmd
Update Initial Community Composition.Rmd
close, but....

---
title: "Initial Biome Composition Analysis"
author: "Timothy Brennhofer"
date: "June 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE, warning = TRUE)

#set working directory
setwd("C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data")
getwd()

#load necessary packages -- "library" or "require"
require(tidyverse)
require(data.table)
require(rmarkdown)
require(ggplot2)
require(dplyr)

#load data
abun_raw <- fread("2016_10_13_run2_otus.csv")
abun <- abun_raw
abun[345,4719] <- 0
abun <- abun %>%
  select(-Run, -Day, -Reads) %>%
  rename(Cage = `Cage#`) %>%
  gather(otu, abun, -SampleID, -Cage, - Abx, -Dose, -Hour, -Cohoused, -Pair) %>%
  group_by(SampleID) %>%
  mutate(abun = abun/sum(abun), otu = as.numeric(otu)) %>%
  mutate(mouseid = paste(Cage, Abx, sep = "_")) %>%
  ungroup()

```

```{r adding taxonomy}

#Load taxonomical data
tax <- fread("C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data/otus_table_joined_taxa.txt")

#Organize tax data into traditional classifications
tax <- tax %>%
  rename(otu = `#OTU ID`, tax = taxonomy) %>%
  mutate(tax = gsub("\\s|.__|\\[|\\]|Other", "", tax)) %>%
  separate(tax, sep = ';', c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), fill = 'right')
tax <- as.data.frame(apply(tax, 2, function(x) ifelse(x == '', 'unclassified', x)), stringsAsFactors = FALSE)
tax$otu <- as.numeric(tax$otu)

#Join taxonomical data and abundance data
abun_tax <- left_join(abun, tax, by = 'otu')

```

```{r intital composition data}

#Filter for inital reads and organize by phyla
abun_tax <- abun_tax %>%
  filter(Hour == 0) %>%
  group_by(mouseid, Phylum)

#Create graph to demonstrate initial compositions
tempz <- abun_tax %>%
  group_by(mouseid, Phylum) %>%
  group_by(mouseid) %>%
  mutate(total = sum(abun))

ggplot(tempz, aes(x = mouseid, y = total, fill = Phylum)) +
  geom_bar()

```

### 2018-06-13 Initial Community Composition.Rmd
Revert "Merge remote-tracking branch 'origin/master'"
This reverts commit 2b74d51665915358512166bea2bd177661db81aa.
Auto stash before revert of "Merge remote-tracking branch 'origin/master'"
Slow progress. Hunting for fixes takes me some time as I am also familiarizing

@@
 ```{r setup, include=FALSE}
-knitr::opts_knit$set(echo = TRUE, warning = TRUE)
+knitr::opts_knit$set(root.dir = "C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data", echo = TRUE, warning = TRUE)
 
 #set working directory
+require(knitr)
+opts_knit$set(root.dir = "C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data")
 setwd("C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data")
@@
 #load data
-abun_raw <- fread("2016_10_13_run2_otus.csv")
+abun_raw <- fread("C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data/2016_10_13_run2_otus.csv")
 abun <- abun_raw
@@
   group_by(mouseid, Phylum) %>%
-  group_by(mouseid) %>%
-  mutate(total = sum(abun))
+  filter(abun > .00001) %>%
+  mutate(percentabun = paste0(round(100 *abun, 4), "%"))
 
-ggplot(tempz, aes(x = mouseid, y = total, fill = Phylum)) +
-  geom_bar()
+tempz <- tempz[order(mouseid, Phylum),]
+
+ggplot(tempz, aes(x = mouseid, y = percentabun, fill = Phylum)) +
+  geom_bar(stat = "identity")
 

### 2018-06-14 Initial Community Composition.Rmd
John comments June 13

@@
 require(knitr)
-opts_knit$set(root.dir = "C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data")
-setwd("C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data")
+opts_knit$set(root.dir = "C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data") #I'm not familiar with this approach
+setwd("C:/Users/Timothy Brennhofer/Desktop/KBS REU/KBS Data") #not sure but the spaces ('white spaces') may be messing it up.
+Or, you may want to add a '\' to the end of the directory.
+But it's fine just to refer to the file directories directly.
 getwd()
@@
 #load necessary packages -- "library" or "require"
-require(tidyverse)
+require(tidyverse) #this includes ggplot and dplyr - no need to load them separately
 require(data.table)
-require(rmarkdown)
-require(ggplot2)
-require(dplyr)
 
@@
 #Filter for inital reads and organize by phyla
+#why alter abun_tax permanently?
 abun_tax <- abun_tax %>%
@@
 
+#Note that this won't do anything; ggplot will plot things alphabetically unless you format the column as a "factor" (google it)
 tempz <- tempz[order(mouseid, Phylum),]
@@
 
+#the problem here is that the paste0() function fills the "percentabun"column with "character" values, as opposed to "number" values.
+Thus, ggplot doesn't know how to plot character-based data in a quantitative way.
+Why not just:
+
+ggplot(abun_tax, aes(x = mouseid, y = abun, fill = Phylum)) +
+  geom_bar(stat = "identity")
+
 ```

### 2018-06-20 mouse_microbiome_john.Rmd
add john scripts

---
title: "mouse_microbiome"
author: "John Guittar"
date: "May 18, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r load data}

# set wd and load packages
wd <- 'C:\\Users\\John\\Documents\\msu\\mouse_microbiome\\'
setwd(wd)
source('custom_functions.R')
loadpax(pkg = c('tidyverse','knitr','data.table'))

# load data
x_raw <- fread('Data\\2018_06_12_run2_otus.csv')
tax <- fread('Data\\otus_table_joined_taxa.txt')

#Organize tax data into traditional classifications
tax <- tax %>%
  rename(otu = `#OTU ID`, tax = taxonomy) %>%
  mutate(tax = gsub("\\s|.__|\\[|\\]|Other", "", tax)) %>%
  separate(tax, sep = ';', c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), fill = 'right')
tax <- as.data.frame(apply(tax, 2, function(x) ifelse(x == '', 'unclassified', x)), stringsAsFactors = FALSE)
tax$otu <- as.numeric(tax$otu)

# Note that, oddly, x[130,4719] is 'NA'. I'm assuming it is zero...?
is.na(x_raw[130,4719])
x_raw[130,4719] <- 0

#remove samples with fewer than 5000 reads
# remove mouse 310 because it dies after hour 120
# create narrow table
# note that there are two samples for Cage 322 at Hour 168. They are merged below
x <- x_raw %>%
  rename(Cage = `Cage#`) %>%
  #mutate(Cage = ifelse(SampleID == 9.192, 319, Cage)) %>%
  filter(Reads > 5000) %>%
  filter(Cage != 310) %>%
  select(-Day, -Run, -Reads) %>%
  gather(otu, abun, -Cage, -Abx, -Hour, -Cohoused, -Dose, -Pair, -SampleID) %>%
  group_by(SampleID, Cage, Abx, Hour, Cohoused, Dose, Pair, otu) %>%
  summarise(abun = sum(abun)) %>%
  group_by(Cage, Hour) %>%
  mutate(abun = abun / sum(abun),
         otu = as.numeric(otu))

#Set plotting theme
th <- theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.background = element_blank())

```

##Percent of total phyla present within single mice, grouped by taxonomic level. Need to add OTU as a final category.

```{r}

j <- tax %>% 
  gather(tax_level, name, -otu) %>% 
  full_join(x, by = 'otu') %>% 
  filter(tax_level != 'Domain') %>% 
  filter(abun > 0)

j <- j %>% 
  mutate(tax_level = factor(tax_level, levels = c('Phylum','Class','Order','Family','Genus','Species'))) %>%
  group_by(tax_level) %>% 
  mutate(total_rich = n_distinct(name)) %>% 
  group_by(Cage, tax_level) %>% 
  mutate(rich = n_distinct(name), val = rich / total_rich)

j %>% 
  ggplot(aes(x = tax_level, y = val, group = Cage)) + 
  geom_line() + 
  expand_limits(y = c(0,1)) +
  th

```

First, I categorize each OTU in each antibiotic treated mouse into *Absent* (never was there), *Extinct* (was there at the end of antibiotic treatment but then disappeared), *Persistent* (was present throughout the experiment), and *Immigrant* (Was absent at the end of AB treatment, but colonized during recover).

The first plot shows the Hour at which immigrants arrive, plotted by the abundance of that immigrant in the control mouse.

```{r}

#I am assuming that OTUs that are absent from a given mouse in the last two samples collected during antibiotic dosing (i.e., at hours 73 and 97) are not in the mouse. This window can be tweaked.

j <- x %>%
  group_by(Cage) %>%
  filter(Hour >= 63 | Abx == 'N') %>%
  group_by(Cage, otu) %>%
  mutate(
    abun0 = sum(abun[Hour <= 97]),
    status0 = if (sum(abun[Hour <= 97]) > 0) 'present' else 'absent',
    status1 = if (sum(abun[Hour > 97]) > 0) 'present' else 'absent',
    status = ifelse(status0 == 'absent' & status1 == 'absent', 'absent',
                    ifelse(status0 == 'absent' & status1 == 'present', 'immigrant',
                           ifelse(status0 == 'present' & status1 == 'absent', 'extinct',
                                  'persistent'))),
    arrival_time = if (status[1] == 'immigrant') min(Hour[Hour > 97 & abun > 0]) else NA) %>%
  ungroup()

tmp <- j %>%
  filter(!is.na(Pair)) %>%
  group_by(Pair, otu) %>%
  mutate(control_abun = ifelse(Abx == 'Y', unique(abun0[Abx == 'N']), 0))

tmp %>% 
  filter(status == 'immigrant') %>%
  #filter(control_abun > 0) %>%
  mutate(
    time_to_arrival = arrival_time - 120,
    control_abun = control_abun * 100 + 0.01) %>%
  ggplot(aes(x = control_abun, y = time_to_arrival)) + 
    geom_jitter(alpha = 0.05, color = 'red') + 
    scale_x_log10() + 
    stat_smooth(color = 'black', method = 'lm') +
    stat_smooth(color = 'black', lty = 3) +
    labs(x = "OTU abundance in cohoused control mouse + 0.1 %", y = 'Hours before OTU arrival in Abx treated mouse') +
    th

```

```{r, fig.height = 6}

#numbers of OTUs, colored by local status
tmp %>% 
  filter(Abx == 'Y' & status %in% c('immigrant','persistent') & abun > 0) %>% 
  mutate(status = as.character(status)) %>%
  ungroup() %>% 
  arrange(status) %>% 
  ggplot(aes(x = factor(Hour), fill = status)) + 
    geom_bar() + 
    facet_wrap(~Pair, ncol = 1) +
    labs(x = "Hour", y = "Number of OTUs") +
    th

#Relative contribution of OTUs, grouped by local status
tmp %>%
  filter(Abx == 'Y' & status %in% c('immigrant','persistent')) %>% 
  mutate(status = as.character(status)) %>%
  ungroup() %>% 
  arrange(status) %>% 
  ggplot(aes(x = factor(Hour), y = abun, fill = status)) + 
    geom_bar(stat = 'identity') + 
    facet_wrap(~Pair, ncol = 1) + 
    labs(x = "Hour", y = "Relative abundances") +
    th

```

### 2018-06-20 custom_functions.R

@@ -1,0 +1,89 @@
+# Custom functions for microbiome analyses. Created Feb 28, 2017.
+
+loadpax <- function(pkg){
+  # (1) checks package installation, (2) installs them if not, then (3) loads them
+    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
+    if (length(new.pkg)) 
+        install.packages(new.pkg, dependencies = TRUE)
+    sapply(pkg, require, character.only = TRUE)
+}
+
+
+grid_arrange_shared_legend <- function(..., ncol = length(list(...)), nrow = 1, position = c("bottom", "right")) {
+  # A function that lets you put multiple plots together and have a single shared legend (from the first plot)
+  # from hadley on the internet...
+  plots <- list(...)
+  position <- match.arg(position)
+  g <- ggplotGrob(plots[[1]] + theme(legend.position = position))$grobs
+  legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
+  lheight <- sum(legend$height)
+  lwidth <- sum(legend$width)
+  gl <- lapply(plots, function(x) x + theme(legend.position="none"))
+  gl <- c(gl, ncol = ncol, nrow = nrow)
+
+  combined <- switch(position,
+                     "bottom" = arrangeGrob(do.call(arrangeGrob, gl),
+                                            legend,
+                                            ncol = 1,
+                                            heights = unit.c(unit(1, "npc") - lheight, lheight)),
+                     "right" = arrangeGrob(do.call(arrangeGrob, gl),
+                                           legend,
+                                           ncol = 2,
+                                           widths = unit.c(unit(1, "npc") - lwidth, lwidth)))
+  grid.newpage()
+  grid.draw(combined)
+
+}
+
+# Multiple plot function
+#
+# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
+# - cols:   Number of columns in layout
+# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
+#
+# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
+# then plot 1 will go in the upper left, 2 will go in the upper right, and
+# 3 will go all the way across the bottom.
+#
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  library(grid)
+
+  # Make a list from the ... arguments and plotlist
[…]

### 2018-07-04 Manuscript_Mouse_Microbiome.Rmd
set up auto word compiling

---
bibliography: C:\\Users\\John\\Documents\\library.bib
csl: C:\\Users\\John\\Documents\\ecology.csl
output:
  bookdown::word_document2:
    reference_docx: C:\\Users\\John\\Documents\\rmd_style_template_single_spaced.docx
    toc: no
  bookdown::html_document2:
    code_folding: hide
    fig_caption: yes
    theme: united
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include = FALSE}

source('C://Users//John//Documents//initialize_manuscript.R')

```

# Disturbance and recovery of mouse gut microbiota in response to antibiotics.

\newline

John Guittar^1,\*^, Aspen Reese^2^, Lawrence David^3^, Timothy Brennhofer^1,4^, Ashley Shade^5,6^, Elena Litchman^1,6,7^

\newline

^1^ Kellogg Biological Station, Michigan State University, 3700 E Gull Lake Dr, Hickory Corners, MI 49060, USA

^2^ Harvard University, Cambridge, MA 02138, USA

^3^ Department of Molecular Genetics and Microbiology, Duke University, Box 3382, Durham, 11 NC 27708, USA

^4^ Grinnell College, 1115 8th Ave, Grinnell, IA 50112, USA

^5^ Department of Microbiology and Molecular Genetics, Department of Plant, Soil and Microbial Sciences, and The Plant Resilience Institute, Michigan State University, East Lansing MI 48840, USA

^6^ Program in Ecology, Evolutionary Biology and Behavior, Michigan State University East Lansing MI 48840, USA

^7^ Department of Integrative Biology, Michigan State University, East Lansing, MI 48824, USA

^\*^ Correspondence author: E-mail: guittarj@msu.edu

\newline

#Abstract

\newline

#Introduction

* We want to improve our understanding of antibiotic disturbance and recovery in the mammalian gut. 
* Most work in this area has focused on shifts in abundance at coarse taxonomic scales. We want to look at disturbance and recovery at the OTU scale, in order to more explicitly examine ecological dynamics.
* One way to better understand ecological mechanisms is to focus on traits. For example, in plant systems, traits related to dispersal ability and rapid growth are often associated with early colonists to distubed areas. Indeed, these traits often form a syndrome characteristic of pioneer species.
* Colonization of the gut after disturbance my be particularly difficult, because it is an abiotic community with strong impediments to dispersal. Spores have been hypothesized to be a life history trait that enhances the probability of dispersal among hosts.

\newline

##*Questions:*

1. Which traits (other than antibiotic resistance) predict OTU tolerance/susceptibility to antibiotics?

2. Which traits predict recovery after antibiotic disturbances?

3. Which traits predict successful immigration from the regional pool during the recovery process?

`r indent` - Are these traits related to dispersal ability and rapid growth, as expected?

\newline

To answer these questions, we will use data from @Reese2018.

\newline

Another way to frame many of these questions is: can OTU traits help to predict the boxes into which OTUs fall in the following diagram:

![](images/winners_losers.png)

`r ref('figure', 'figOTUFates')`

(ref:figOTUFates) Figure `r refs$figOTUFates`

**(ref:figOTUFates):** The potential fates and fortunes of OTUs in mice treated with antibiotics. Positive outcomes are shaded light green, whereas negative outcomes are shaded light red.

\newline

# Methods

\newline

# Results

```{r, include=FALSE}

# load packages and custom functions
wd <- 'C:\\Users\\John\\Documents\\msu\\mouse_microbiome\\'
source(paste0(wd, 'custom_functions.R'))
loadpax(pkg = c('tidyverse','knitr','data.table','vegan', 'ape'))

#set directory for printed images/figures
image_dir <- paste0(wd, 'images\\')

# load data
x_raw <- fread('C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\2018_06_12_run2_otus.csv')
tax <- fread('C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\otus_table_joined_taxa.txt')

#Organize tax data into traditional classifications
tax <- tax %>%
  rename(otu = `#OTU ID`, tax = taxonomy) %>%
  mutate(tax = gsub("\\s|.__|\\[|\\]|Other", "", tax)) %>%
  separate(tax, sep = ';', c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), fill = 'right')
tax <- as.data.frame(apply(tax, 2, function(x) ifelse(x == '', 'unclassified', x)), stringsAsFactors = FALSE)
tax$otu <- as.numeric(tax$otu)

# Note that, oddly, x[130,4719] is 'NA'. I'm assuming it is zero...?
x <- as.data.frame(x_raw)
is.na(x[130,4719])
x[130,4719] <- 0

#remove samples with fewer than 5000 reads.
x <- x[rowSums(x[c(11:ncol(x))]) >= 5000, ]

#remove cage 310 because mouse dies
#x <- x[x$`Cage#` != 310, ]

#ensure all OTUs occur at least once after removing samples
x <- bind_cols(x[, c(1:10)], x[, c(11:ncol(x))[colSums(x[,c(11:ncol(x))]) > 0]])

#save unrarified version of data?
x_unrarified <- x

#rarefy to 5000
tmp <- x[,c(11:ncol(x))]
tmp <- tmp[rowSums(tmp) >= 5000,]
tmp <- rrarefy(tmp, 5000)
x <- bind_cols(x[, c(1:10)], as.data.frame(tmp))

#create wide table for distance calculations
x <- x %>%
  mutate(Mouse = floor(SampleID)) %>%
  select(-Day, -Run, -Reads, -`Cage#`, -SampleID)

#fix data entry problem1
filt <- x$Mouse == 9 & x$Hour == 168
x$Cohoused[filt] <- 'Y'
x$Pair[filt] <- 2
x$Abx[filt] <- 'N'
x$Dose[filt] <- 'No'

#fix data entry problem2
filt <- x$Mouse == 7 & x$Hour == 168
x$Cohoused[filt] <- 'Y'
x$Pair[filt] <- 4
x$Abx[filt] <- 'Y'
x$Dose[filt] <- 'Recovery'

#save wide table for later
x_wide <- x

#cleanup, make narrow table
# I could relativize to a percent here, but why?
x <- x %>%
  gather(otu, abun, -Mouse, -Abx, -Hour, -Cohoused, -Dose, -Pair) %>%
  group_by(Mouse, Abx, Hour, Cohoused, Dose, Pair, otu) %>%
  summarise(abun = sum(abun)) %>%
  mutate(otu = as.numeric(otu))

#set up phyloseq object for UniFrac distances.
#Want to do it but can't until I make a tree...

#Set plotting theme
th <- theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.background = element_blank())

```

```{r, include = FALSE}

#####
#takes too long
#ord <- metaMDS(x_wide, k = 3, trymax = 1000, noshare = 0.2)
#####

myp <- select(x_wide, -Mouse, -Abx, -Dose, -Hour, -Cohoused, -Pair)
myp <- pcoa(vegdist(myp))

j <- data.frame(
  x_wide[, c('Mouse','Abx','Hour','Pair')], 
  PCoA1 = myp$vectors[,1],
  PCoA2 = myp$vectors[,2]) %>%
  group_by(Mouse) %>%
  mutate(
  label = ifelse(Abx == 'Y' & (Hour == min(Hour) | Hour == 120 | Hour == max(Hour)), Hour, NA),
  Abx = factor(ifelse(Abx == 'Y','Abx', 'No abx'), levels = c('No abx', 'Abx')))

p1 <- j %>% 
  filter(!is.na(Pair)) %>%
  mutate(Pair = paste('Cage', Pair)) %>%
  ggplot(aes(x = PCoA1, y = PCoA2, color = Abx, fill = Abx)) + 
  geom_point(aes(alpha = Hour), shape = 21, size = 4) + 
  geom_point(shape = 21, size = 4, fill = NA) + 
  geom_path() + 
  geom_text(aes(label = label), color = 'black', na.rm=TRUE) +
  facet_grid(Pair~Abx) + th

plot_it(p1, name = 'figPCoAPaired', dir = image_dir, width = 7, height = 8)

```

(ref:figPCoAPaired) Figure `r refs$figPCoAPaired`

`r pagebreak` ![](images\\figPCoAPaired.png)

**(ref:figPCoAPaired):** Figure legend.

```{r, include = FALSE}

p2 <- j %>% 
  filter(is.na(Pair)) %>%
  group_by(Abx) %>%
  mutate(Pair2 = as.numeric(as.factor(Mouse))) %>%
  ggplot(aes(x = PCoA1, y = PCoA2, color = Abx, fill = Abx)) + 
  geom_point(aes(alpha = Hour), shape = 21, size = 4) + 
  geom_point(shape = 21, size = 4, fill = NA) + 
  geom_path() + 
  geom_text(aes(label = label), color = 'black', na.rm=TRUE) +
  facet_grid(Pair2~Abx) + th

plot_it(p2, name = 'figPCoASolitary', dir = image_dir, width = 7, height = 8)

```

(ref:figPCoASolitary) Figure `r refs$figPCoASolitary`

`r pagebreak` ![](images\\figPCoASolitary.png)

**(ref:figPCoASolitary):** Figure legend.

```{r, include = FALSE}

j <- tax %>% 
  gather(tax_level, name, -otu) %>% 
  full_join(x, by = 'otu') %>% 
  filter(tax_level != 'Domain') %>% 
  filter(abun > 0)

j <- j %>% 
  mutate(tax_level = factor(tax_level, levels = c('Phylum','Class','Order','Family','Genus','Species'))) %>%
  group_by(tax_level) %>% 
  mutate(total_rich = n_distinct(name)) %>% 
  group_by(Mouse, tax_level) %>% 
  mutate(rich = n_distinct(name), val = rich / total_rich)

p1 <- j %>% 
  ggplot(aes(x = tax_level, y = val, group = Mouse)) + 
  geom_line() + 
  expand_limits(y = c(0,1)) +
  th

plot_it(p1, name = 'figTaxaSharedAcrossMice', dir = image_dir, width = 5, height = 4)

```

(ref:figTaxaSharedAcrossMice) Figure `r refs$figTaxaSharedAcrossMice`

`r pagebreak` ![](images\\figTaxaSharedAcrossMice.png)

**(ref:figTaxaSharedAcrossMice):** Percent of total phyla present within single mice, grouped by taxonomic level. Need to add OTU as a final category. First, I categorize each OTU in each antibiotic treated mouse into *Absent* (never was there), *Extinct* (was there at the end of antibiotic treatment but then disappeared), *Persistent* (was present throughout the experiment), and *Immigrant* (Was absent at the end of AB treatment, but colonized during recover).

The first plot shows the Hour at which immigrants arrive, plotted by the abundance of that immigrant in the control mouse.

```{r, include = FALSE}

#I am assuming that OTUs that are absent from a given mouse in the last two samples collected during antibiotic dosing (i.e., at hours 73 and 97) are not in the mouse. This window can be tweaked.

j <- x %>%
  group_by(Mouse) %>%
  filter(Hour >= 63 | Abx == 'N') %>%
  group_by(Mouse, otu) %>%
  mutate(
    abun0 = sum(abun[Hour <= 97]),
    status0 = if (sum(abun[Hour <= 97]) > 0) 'present' else 'absent',
    status1 = if (sum(abun[Hour > 97]) > 0) 'present' else 'absent',
    status = ifelse(status0 == 'absent' & status1 == 'absent', 'absent',
                    ifelse(status0 == 'absent' & status1 == 'present', 'immigrant',
                           ifelse(status0 == 'present' & status1 == 'absent', 'extinct',
                                  'persistent'))),
    arrival_time = if (status[1] == 'immigrant') min(Hour[Hour > 97 & abun > 0]) else NA) %>%
  ungroup()

tmp <- j %>%
  filter(!is.na(Pair)) %>%
  group_by(Pair, otu) %>%
  mutate(control_abun = ifelse(Abx == 'Y', unique(abun0[Abx == 'N']), 0))

p <- tmp %>% 
  filter(status == 'immigrant') %>%
  #filter(control_abun > 0) %>%
  mutate(
    time_to_arrival = arrival_time - 120,
    control_abun = control_abun * 100 + 0.01) %>%
  ggplot(aes(x = control_abun, y = time_to_arrival)) + 
    geom_jitter(alpha = 0.05, color = 'red') + 
    scale_x_log10() + 
    stat_smooth(color = 'black', method = 'lm') +
    stat_smooth(color = 'black', lty = 3) +
    labs(x = "OTU abundance in cohoused control mouse + 0.1 %", y = 'Hours before OTU arrival in Abx treated mouse') +
    th

plot_it(p, name = 'figPropPressure', dir = image_dir, width = 6.5, height = 5)

```

(ref:figPropPressure) Figure `r refs$figPropPressure`

`r pagebreak` ![](images\\figPropPressure.png)

**(ref:figPropPressure):** Figure legend.

```{r, include = FALSE}

#numbers of OTUs, colored by local status
tmp %>% 
  filter(Abx == 'Y' & status %in% c('immigrant','persistent') & abun > 0) %>% 
  mutate(status = as.character(status)) %>%
  ungroup() %>% 
  arrange(status) %>% 
  ggplot(aes(x = factor(Hour), fill = status)) + 
    geom_bar() + 
    facet_wrap(~Pair, ncol = 1) +
    labs(x = "Hour", y = "Number of OTUs") +
    th

#Relative contribution of OTUs, grouped by local status
p <- tmp %>%
  filter(Abx == 'Y' & status %in% c('immigrant','persistent')) %>% 
  mutate(status = as.character(status)) %>%
  ungroup() %>% 
  arrange(status) %>% 
  ggplot(aes(x = factor(Hour), y = abun, fill = status)) + 
    geom_bar(stat = 'identity') + 
    facet_wrap(~Pair, ncol = 1) + 
    labs(x = "Hour", y = "Relative abundances") +
    th

plot_it(p, name = 'figImmAbunTime', dir = image_dir, height = 7, width = 5)

```

(ref:figImmAbunTime) Figure `r refs$figImmAbunTime`

`r pagebreak` ![](images\\figImmAbunTime.png)

**(ref:figImmAbunTime):** Figure legend.

```{r, include = FALSE}

st <- x %>%
  group_by(otu, Day = floor(Hour/24)) %>%
  filter(sum(abun[Abx == 'Y'] > 0) & sum(abun[Abx == 'N'] > 0)) %>%
  do(mod = t.test(abun~Abx, data = .)) %>%
  mutate(
    pval = mod$p.value,
    df = mod$parameter,
    tval = mod$statistic,
    mean_No_Abx = mod$estimate[[1]],
    mean_Yes_Abx = mod$estimate[[2]],
    diff = mean_No_Abx - mean_Yes_Abx)

j <- ungroup(st) %>% 
  group_by(otu) %>% 
  summarise(sigs = sum(pval < 0.05), sign = sum(diff)) %>% 
  arrange(desc(abs(sign))) %>%
  do(data.frame(winners = 
    head(.$otu[.$sign < 0]),
    losers = head(.$otu[.$sign > 0]))) %>%
  gather(status, otu)

p <- x %>%
  left_join(tax) %>%
  filter(otu %in% j$otu) %>%
  mutate(label = paste(Family, 'OTU', as.numeric(as.factor(otu)))) %>%
  left_join(j) %>%
  ggplot(aes(x = Hour, y = abun + 1, color = Abx)) + 
  annotate(geom = 'rect', xmin = 1, xmax = 120, ymin = 0, ymax = Inf, alpha = 0.3, fill = 'red') + 
  geom_point() + 
  stat_smooth() +
  #geom_text(label = "Antibiotics", x = 60, aes(y = max(abun))) +
  #geom_text(label = "Recovery", x = 216, aes(y = max(abun))) +
  scale_y_log10() +
  scale_color_manual(values = c('black','red')) +
  facet_wrap(~label) + 
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    panel.background = element_blank(),
    legend.position = 'bottom')

plot_it(p, name = 'figMostAffectedOTUs', dir = image_dir, width = 9, height = 8)

```

(ref:figMostAffectedOTUs) Figure `r refs$figMostAffectedOTUs`

`r pagebreak` ![](images\\figImmAbunTime.png)

**(ref:figMostAffectedOTUs):** Figure legend.

`r pagebreak` Works Cited

### 2018-07-06 Manuscript_Mouse_Microbiome.Rmd
more figs

@@
 
-```{r setup, include = FALSE}
+```{r, include=FALSE}
 
+#initialize manuscript formatting script
 source('C://Users//John//Documents//initialize_manuscript.R')
@@
 
+`r pagebreak`
+
+![](images/setup_Reese2018.png)
+
+`r ref('figure', 'figReese2018Setup')`
+
+(ref:figReese2018Setup) Figure `r refs$figReese2018Setup`
+
+**(ref:figReese2018Setup):** A figure drawn from @Reese2018 showing the experimental setup and data collection schedule.
+
+`r pagebreak`
+
 \newline
@@
 
-#fix data entry problem1
-filt <- x$Mouse == 9 & x$Hour == 168
-x$Cohoused[filt] <- 'Y'
-x$Pair[filt] <- 2
-x$Abx[filt] <- 'N'
-x$Dose[filt] <- 'No'
-
-#fix data entry problem2
-filt <- x$Mouse == 7 & x$Hour == 168
-x$Cohoused[filt] <- 'Y'
-x$Pair[filt] <- 4
-x$Abx[filt] <- 'Y'
-x$Dose[filt] <- 'Recovery'
+#fix data entry problems
+x$Cohoused[x$Mouse == 3 & x$Hour == 168] <- 'N'
+x$Cohoused[x$Mouse == 6 & x$Hour == 168] <- 'N'
+x$Cohoused[x$Mouse == 7 & x$Hour == 168] <- 'Y'
+x$Pair[x$Mouse == 7 & x$Hour == 168] <- 4
+x$Abx[x$Mouse == 7 & x$Hour == 168] <- 'Y'
+x$Dose[x$Mouse == 7 & x$Hour == 168] <- 'Recovery'
+x$Cohoused[x$Mouse == 9 & x$Hour == 168] <- 'Y'
+x$Pair[x$Mouse == 9 & x$Hour == 168] <- 2
+x$Abx[x$Mouse == 9 & x$Hour == 168] <- 'N'
+x$Dose[x$Mouse == 9 & x$Hour == 168] <- 'No'
+x$Cohoused[x$Mouse == 12 & x$Hour == 168] <- 'N'
+x$Cohoused[x$Mouse == 16 & x$Hour == 168] <- 'N'
 
@@
   geom_text(aes(label = label), color = 'black', na.rm=TRUE) +
-  facet_grid(Pair~Abx) + th
+  scale_color_manual(values = c('black','red')) +
+  scale_fill_manual(values = c('black','red')) +
+  facet_grid(Pair~Abx) + 
+  th
 
 plot_it(p1, name = 'figPCoAPaired', dir = image_dir, width = 7, height = 8)
+
+p1
 
@@
 
-**(ref:figPCoAPaired):** Figure legend.
+**(ref:figPCoAPaired):** PCoA of mouse microbiomes over time, during and after 120 hours of antibiotic treatment.
+Control mice (left) and treated mice (right) were cohoused after treatment (hour 120) for the duration of the experiment.
 
@@
   geom_text(aes(label = label), color = 'black', na.rm=TRUE) +
+  scale_color_manual(values = c('black','red')) +
+  scale_fill_manual(values = c('black','red')) +
   facet_grid(Pair2~Abx) + th
@@
 plot_it(p2, name = 'figPCoASolitary', dir = image_dir, width = 7, height = 8)
+
+p2
 
@@
 
-**(ref:figPCoASolitary):** Figure legend.
+**(ref:figPCoASolitary):** PCoA of mouse microbiomes over time, during and after 120 hours of antibiotic treatment.
+Control mice (left) and treated mice (right) were not cohoused at any point during the experiment.
 
@@
   full_join(x, by = 'otu') %>% 
+  filter(Hour <= 120) %>%
   filter(tax_level != 'Domain') %>% 
@@
   group_by(Mouse, tax_level) %>% 
-  mutate(rich = n_distinct(name), val = rich / total_rich)
+  mutate(rich = n_distinct(name), val = rich / total_rich,
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'))
 
 p1 <- j %>% 
-  ggplot(aes(x = tax_level, y = val, group = Mouse)) + 
-  geom_line() + 
+  ggplot(aes(x = tax_level, y = val, group = Mouse, color = Abx)) + 
+  #geom_line() + 
+  stat_summary(fun.data = "mean_sdl", fun.args = list(mult=1), aes(group = Abx), position = position_dodge(width = 0.2)) +
+  scale_color_manual(values = c('black','red')) +
   expand_limits(y = c(0,1)) +
+  labs(x = 'Taxonomic level', y = 'Percent of all Phyla present across all mice')
   th
@@
 plot_it(p1, name = 'figTaxaSharedAcrossMice', dir = image_dir, width = 5, height = 4)
+
+p1
 
@@
 
-**(ref:figTaxaSharedAcrossMice):** Percent of total phyla present within single mice, grouped by taxonomic level.
-Need to add OTU as a final category.
-First, I categorize each OTU in each antibiotic treated mouse into *Absent* (never was there), *Extinct* (was there at the end of antibiotic treatment but then disappeared), *Persistent* (was present throughout the experiment), and *Immigrant* (Was absent at the end of AB treatment, but colonized during recover).
-
-The first plot shows the Hour at which immigrants arrive, plotted by the abundance of that immigrant in the control mouse.
+**(ref:figTaxaSharedAcrossMice):** Percent of phyla observed across all mice present within single mice, assessed at increasingly specific taxonomic levels.
 
@@
 
-#I am assuming that OTUs that are absent from a given mouse in the last two samples collected during antibiotic dosing (i.e., at hours 73 and 97) are not in the mouse. This window can be tweaked.
+j <- x %>%
+  group_by(Mouse, Hour) %>%
+  mutate(rich = length(unique(otu[abun > 0]))) %>%
+  group_by(Mouse, Hour, Abx, Cohoused, Dose, Pair, rich) %>%
+  do(shannon = diversity(spread(select(., otu, abun), otu, abun))) %>%
+  mutate(shannon_val = shannon[[1]])
+
+#must fix screwy data!!! (when I paste mouse, abx, cohoused, there are some unexpected combinations)
+j <- j %>%
+  group_by(Mouse, Abx, Cohoused) %>%
+  filter(length(shannon_val) > 1) %>%
+  ungroup() %>%
+  mutate(
+    evenness = shannon_val / log(rich),
+    Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'),
+    Mouse = ifelse(Cohoused == 'Y', paste(Abx, "Cohoused", Mouse), paste(Abx, 'Solitary', Mouse)))
+
+p <- j %>%
+  ggplot(aes(x = Hour, y = shannon_val, color = Abx)) +
+    geom_point() +
+    stat_smooth() +
+    scale_color_manual(values = c('red','black')) +
+    geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data = j %>% group_by(Mouse) %>% filter(row_number() == 1 & Abx == 'Abx')) + 
+    facet_wrap(~Mouse, ncol = 3) +
+    labs(x = 'Hour', y = 'Shannon diversity') +
+    th
+
+plot_it(p, name = 'figShannonOverTime', dir = image_dir, width = 6, height = 8)
+
+p
+
+```
+
+(ref:figShannonOverTime) Figure `r refs$figShannonOverTime`
+
+`r pagebreak` ![](images\\figShannonOverTime.png)
+
+**(ref:figShannonOverTime):** Shannon diversity over time within mice.
+
+```{r, include = FALSE}
+
+p <- j %>%
+  ggplot(aes(x = Hour, y = rich, color = Abx)) +
+    geom_point() +
+    stat_smooth() +
+    scale_color_manual(values = c('red','black')) +
+    geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data = j %>% group_by(Mouse) %>% filter(row_number() == 1 & Abx == 'Abx')) + 
+    facet_wrap(~Mouse, ncol = 3) +
+    labs(x = 'Hour', y = 'OTU richness') +
+    th
+
+plot_it(p, name = 'figRichOverTime', dir = image_dir, width = 6, height = 8)
+
+p
+
+```
+
+(ref:figRichOverTime) Figure `r refs$figRichOverTime`
+
+`r pagebreak` ![](images\\figRichOverTime.png)
+
+**(ref:figRichOverTime):** OTU richness over time within mice.
+
+```{r, include = FALSE}
+
+p <- j %>%
+  ggplot(aes(x = Hour, y = evenness, color = Abx)) +
+    geom_point() +
+    stat_smooth() +
+    scale_color_manual(values = c('red','black')) +
+    geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data = j %>% group_by(Mouse) %>% filter(row_number() == 1 & Abx == 'Abx')) + 
+    facet_wrap(~Mouse, ncol = 3) +
+    labs(x = 'Hour', y = 'Community evenness') +
+    th
+
+plot_it(p, name = 'figEvennessOverTime', dir = image_dir, width = 6, height = 8)
+
+p
+
+```
+
+(ref:figEvennessOverTime) Figure `r refs$figEvennessOverTime`
+
+`r pagebreak` ![](images\\figEvennessOverTime.png)
+
+**(ref:figEvennessOverTime):** OTU community evenness (i.e., equitability, i.e., Shannon diversity H / log(OTU richness))
+
+```{r, include = FALSE}
+
+#I am assuming that OTUs that are absent from a given mouse in the last three samples collected during antibiotic dosing (i.e., at hours 63, 73 and 97) are not in the mouse.
+This window can be tweaked.
+
+#set the window. The edges are included.
+window0 <- 63 
+window1 <- 120
 
 j <- x %>%
-  group_by(Mouse) %>%
-  filter(Hour >= 63 | Abx == 'N') %>%
-  group_by(Mouse, otu) %>%
+  group_by(Mouse, Abx, Pair, otu) %>%
+  summarise(
+    abunPre = if (unique(Abx) == 'Y') sum(abun[Hour <= window0]) else NA,
+    abun0 = if (unique(Abx) == 'Y') sum(abun[Hour >= window0 & Hour <= window1]) else sum(abun[Hour < window1]),
+    abun1 = sum(abun[Hour > window1]),
+    statusAB = if (!is.na(abunPre) & abunPre == 0 & abun0 == 0) 'absent' else 'persistent',
+    statusAB = if (!is.na(abunPre) & abunPre > 0 & abun0 == 0) 'extinct' else statusAB,
+    ETD = if (statusAB == 'extinct') max(Hour[Hour <= window0 & abun > 0]) else NA,
+    ETD = if (statusAB == 'persistent') Inf else ETD,
+    status0 = if (abun0 > 0) 'present' else 'absent',
+    status1 = if (sum(abun[Hour > window1] > 0)) 'present' else 'absent',
+    status = if (status0 == 'absent' & status1 == 'absent') 'absent' else 'persistent',
+    status = if (status0 == 'absent' & status1 == 'present') 'immigrant' else status,
+    status = if (status0 == 'present' & status1 == 'absent') 'extinct' else status,
+    ETA = if (status == 'immigrant') min(Hour[Hour > 97 & abun > 0]) else NA) %>%
+  group_by(Pair, otu) %>%
   mutate(
-    abun0 = sum(abun[Hour <= 97]),
-    status0 = if (sum(abun[Hour <= 97]) > 0) 'present' else 'absent',
-    status1 = if (sum(abun[Hour > 97]) > 0) 'present' else 'absent',
-    status = ifelse(status0 == 'absent' & status1 == 'absent', 'absent',
-                    ifelse(status0 == 'absent' & status1 == 'present', 'immigrant',
-                           ifelse(status0 == 'present' & status1 == 'absent', 'extinct',
-                                  'persistent'))),
-    arrival_time = if (status[1] == 'immigrant') min(Hour[Hour > 97 & abun > 0]) else NA) %>%
-  ungroup()
+    abun_pair = ifelse(!is.na(Pair) & Abx == 'Y', abun0[Abx == 'N'], NA),
+    abun_pair = ifelse(!is.na(Pair) & Abx == 'N', abun0[Abx == 'Y'], abun_pair))
 
-tmp <- j %>%
-  filter(!is.na(Pair)) %>%
+j <- x %>%
+  group_by(Mouse, Abx, Pair, otu) %>%
+  summarise(
+    abun0 = sum(abun[Hour < window0]),
+    abun1 = sum(abun[Hour >= window0 & Hour <= window1]),
+    abun01 = abun0 + abun1,
+    abun2 = sum(abun[Hour > window1]),
+    status_treatment = if (abun01 == 0) 'absent' 
+                         else if (abun0 > 0 & abun1 == 0) 'extinct' 
+                           else 'persistent',
+    ETD = if (status_treatment == 'extinct') max(Hour[Hour <= window0 & abun > 0]) 
+            else if (status_treatment == 'persistent') Inf 
+              else NA,
+    status_recovery = if (abun01 == 0 & abun2 == 0) 'absent' 
+               else if (abun1 == 0 & abun2 > 0) 'immigrant'
+                 else if (abun01 > 0 & abun2 == 0) 'extinct'
+                   else 'persistent',
+    immigrant_type = if (status_recovery != 'immigrant') 'Not an immigrant'
+                       else if (abun01 > 0) 'recovered'
+                         else 'new colonist',
+    ETA = if (status_recovery == 'immigrant') min(Hour[Hour > 97 & abun > 0]) 
+            else NA) %>%
   group_by(Pair, otu) %>%
-  mutate(control_abun = ifelse(Abx == 'Y', unique(abun0[Abx == 'N']), 0))
+  mutate(
+    abun_pair = ifelse(!is.na(Pair) & Abx == 'Y', abun01[Abx == 'N'], NA),
+    abun_pair = ifelse(!is.na(Pair) & Abx == 'N', abun01[Abx == 'Y'], abun_pair))
 
-p <- tmp %>% 
-  filter(status == 'immigrant') %>%
-  #filter(control_abun > 0) %>%
-  mutate(
-    time_to_arrival = arrival_time - 120,
-    control_abun = control_abun * 100 + 0.01) %>%
-  ggplot(aes(x = control_abun, y = time_to_arrival)) + 
-    geom_jitter(alpha = 0.05, color = 'red') + 
+p <- j %>% 
+  filter(status_recovery == 'immigrant' & Abx == 'Y' & abun_pair > 0) %>%
+  mutate(time_to_arrival = ETA - 120) %>%
+  ggplot(aes(x = abun_pair, y = time_to_arrival)) + 
+    geom_jitter(alpha = 0.25, color = 'red', height = 5, width = 0) + 
     scale_x_log10() + 
@@
     stat_smooth(color = 'black', lty = 3) +
-    labs(x = "OTU abundance in cohoused control mouse + 0.1 %", y = 'Hours before OTU arrival in Abx treated mouse') +
+    labs(x = "OTU relative abundance in cohoused control mouse", y = 'Hours before OTU arrival in Abx treated mouse') +
     th
@@
 plot_it(p, name = 'figPropPressure', dir = image_dir, width = 6.5, height = 5)
+
+p
 
@@
 
-**(ref:figPropPressure):** Figure legend.
+**(ref:figPropPressure):** A test of the potential role of mass effects.
+The hypothesis is that, all else being equal, the abundance of a taxon in the regional pool (i.e., the cohoused mouse) is going to correlate positively with time to arrival (i.e., successful colonization) in the treated mouse.
+This hypothesis is supported.
+However, we can't be sure if the pattern is because the taxa that are more abundant taxa in the control mouse are simply better suited for the gut environment, and thus are faster to colonize the antibiotic-treated mouse.
+So, the analysis has it's limitations, but I still think it's worth noting.
 
@@
 
-#numbers of OTUs, colored by local status
-tmp %>% 
-  filter(Abx == 'Y' & status %in% c('immigrant','persistent') & abun > 0) %>% 
-  mutate(status = as.character(status)) %>%
+p <- x %>%
   ungroup() %>%
-  arrange(status) %>% 
-  ggplot(aes(x = factor(Hour), fill = status)) + 
-    geom_bar() + 
-    facet_wrap(~Pair, ncol = 1) +
-    labs(x = "Hour", y = "Number of OTUs") +
-    th
+  left_join(j[, c('Mouse','otu','status_recovery','immigrant_type')], by = c('Mouse','otu')) %>%
+  filter(!is.na(Pair) & abun > 0 & status_recovery != 'absent') %>%
+  arrange(status_recovery) %>%
+  mutate(status_recovery = ifelse(status_recovery == 'immigrant', immigrant_type, status_recovery)) %>%
+  mutate(OTU_status = c('Extinct by end of Abx treatment (hour 120 at the latest); does not return during recovery period', 
+                        'New colonist, arrived during recovery period (after hour 120)',
+                        'Present throughout experiment',
+                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
+         Pair = paste('Pair', Pair),
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx')) %>%
+  ggplot(aes(x = as.factor(Hour), y = abun, fill = OTU_status)) +
+    geom_bar(stat='identity', alpha = 0.8) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    facet_grid(Abx~Pair) +
+    scale_fill_manual(name = '', values = c('#a6611a','#dfc27d','#80cdc1','#018571')) +
+    #scale_fill_discrete(name = '') +
+    labs(x = 'Hour: {0,4,8,12,16,20,23,25,39,49,63,73,97,120,144,168,192,216,240,312}', y = 'Rarefied abundance') +
+    guides(fill=guide_legend(nrow=4,byrow=TRUE, position = 'bottom')) +
+    theme_bw() +
+    theme(
+      panel.grid.minor = element_blank(),
+      panel.grid.major = element_blank(),
+      panel.background = element_blank(),
+      axis.text.x=element_blank(),
+      legend.position = 'bottom')
 
-#Relative contribution of OTUs, grouped by local status
-p <- tmp %>%
-  filter(Abx == 'Y' & status %in% c('immigrant','persistent')) %>% 
-  mutate(status = as.character(status)) %>%
-  ungroup() %>% 
-  arrange(status) %>% 
-  ggplot(aes(x = factor(Hour), y = abun, fill = status)) + 
-    geom_bar(stat = 'identity') + 
-    facet_wrap(~Pair, ncol = 1) + 
-    labs(x = "Hour", y = "Relative abundances") +
-    th
+plot_it(p, name = 'figOTUAbunByAbxStatus', dir = image_dir, width = 6.5, height = 5)
 
-plot_it(p, name = 'figImmAbunTime', dir = image_dir, height = 7, width = 5)
+p
 
@@
 
-(ref:figImmAbunTime) Figure `r refs$figImmAbunTime`
+(ref:figOTUAbunByAbxStatus) Figure `r refs$figOTUAbunByAbxStatus`
 
-`r pagebreak` ![](images\\figImmAbunTime.png)
+`r pagebreak` ![](images\\figOTUAbunByAbxStatus.png)
 
-**(ref:figImmAbunTime):** Figure legend.
+**(ref:figOTUAbunByAbxStatus):** Rarefied OTU abundances, colored by OTU status. This figure is to compare mice treated by antibiotics to untreated mice.
+
+```{r, include = FALSE}
+
+x %>%
+  ungroup() %>%
+  left_join(j[, c('Mouse','otu','status_recovery','immigrant_type')], by = c('Mouse','otu')) %>%
+  filter(Abx == 'Y' & abun > 0 & status_recovery != 'absent' & Mouse != 1) %>%
+  arrange(status_recovery) %>%
+  mutate(status_recovery = ifelse(status_recovery == 'immigrant', immigrant_type, status_recovery)) %>%
+  mutate(OTU_status = c('Extinct by end of Abx treatment (hour 120 at the latest); does not return during recovery period', 
+                        'New colonist, arrived during recovery period (after hour 120)',
+                        'Present throughout experiment',
+                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
+         Mouse = paste(ifelse(Cohoused == 'Y', 'Cohoused', 'Not cohoused'), "- Mouse", Mouse),
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx')) %>%
+  ggplot(aes(x = as.factor(Hour), y = abun, fill = OTU_status)) +
+    geom_bar(stat='identity', alpha = 0.8) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    facet_wrap(~Mouse, ncol = 5) +
+    scale_fill_manual(name = '', values = c('#a6611a','#dfc27d','#80cdc1','#018571')) +
+    #scale_fill_discrete(name = '') +
+    labs(x = 'Hour: {0,4,8,12,16,20,23,25,39,49,63,73,97,120,144,168,192,216,240,312}', y = 'Rarefied abundance') +
+    guides(fill=guide_legend(nrow=4,byrow=TRUE, position = 'bottom')) +
+    theme_bw() +
+    theme(
+      panel.grid.minor = element_blank(),
+      panel.grid.major = element_blank(),
+      panel.background = element_blank(),
+      axis.text.x=element_blank(),
+      legend.position = 'bottom')
+
+plot_it(p, name = 'figOTUAbunByCageStatus', dir = image_dir, width = 6.5, height = 5)
+
+p
+
+```
+
+(ref:figOTUAbunByCageStatus) Figure `r refs$figOTUAbunByCageStatus`
+
+`r pagebreak` ![](images\\figOTUAbunByCageStatus.png)
+
+**(ref:figOTUAbunByCageStatus):** Rarefied OTU abundances, colored by OTU status.
+This figure is to compare mice in cohousing environments during the recovery process, with mice caged alone during the recovery process.
+
+```{r, include = FALSE}
+
+x %>%
+  ungroup() %>%
+  left_join(j[, c('Mouse','otu','status_recovery','immigrant_type')], by = c('Mouse','otu')) %>%
+  filter(!is.na(Pair) & abun > 0 & status_recovery != 'absent') %>%
+  arrange(status_recovery) %>%
+  mutate(status_recovery = ifelse(status_recovery == 'immigrant', immigrant_type, status_recovery)) %>%
+  mutate(OTU_status = c('Extinct by end of Abx treatment (hour 120 at the latest); does not return during recovery period', 
+                        'New colonist, arrived during recovery period (after hour 120)',
+                        'Present throughout experiment',
+                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
+         Pair = paste('Pair', Pair),
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx')) %>%
+  ggplot(aes(x = as.factor(Hour), fill = OTU_status)) +
+    geom_bar(alpha = 0.8) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    facet_grid(Abx~Pair) +
+    scale_fill_manual(name = '', values = c('#a6611a','#dfc27d','#80cdc1','#018571')) +
+    #scale_fill_discrete(name = '') +
+    labs(x = 'Hour: {0,4,8,12,16,20,23,25,39,49,63,73,97,120,144,168,192,216,240,312}', y = 'OTU richness') +
+    guides(fill=guide_legend(nrow=4,byrow=TRUE, position = 'bottom')) +
+    theme_bw() +
+    theme(
+      panel.grid.minor = element_blank(),
+      panel.grid.major = element_blank(),
+      panel.background = element_blank(),
+      axis.text.x=element_blank(),
+      legend.position = 'bottom')
+
+plot_it(p, name = 'figOTURichByAbxStatus', dir = image_dir, width = 6.5, height = 5)
+
+p
+
+```
+
+(ref:figOTURichByAbxStatus) Figure `r refs$figOTURichByAbxStatus`
+
+`r pagebreak` ![](images\\figOTURichByAbxStatus.png)
+
+**(ref:figOTURichByAbxStatus):** OTU richness from communities rarefied to 5000 seqeunces, colored by OTU status.
+This figure is to compare mice in cohousing environments during the recovery process, with mice caged alone during the recovery process.
+
+```{r, include = FALSE}
+
+x %>%
+  ungroup() %>%
+  left_join(j[, c('Mouse','otu','status_recovery','immigrant_type')], by = c('Mouse','otu')) %>%
+  filter(Abx == 'Y' & abun > 0 & status_recovery != 'absent' & Mouse != 1) %>%
+  arrange(status_recovery) %>%
+  mutate(status_recovery = ifelse(status_recovery == 'immigrant', immigrant_type, status_recovery)) %>%
+  mutate(OTU_status = c('Extinct by end of Abx treatment (hour 120 at the latest); does not return during recovery period', 
+                        'New colonist, arrived during recovery period (after hour 120)',
+                        'Present throughout experiment',
+                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
+         Mouse = paste(ifelse(Cohoused == 'Y', 'Cohoused', 'Not cohoused'), "- Mouse", Mouse),
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx')) %>%
+  ggplot(aes(x = as.factor(Hour), fill = OTU_status)) +
+    geom_bar(alpha = 0.8) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    facet_wrap(~Mouse, ncol = 5) +
+    scale_fill_manual(name = '', values = c('#a6611a','#dfc27d','#80cdc1','#018571')) +
+    #scale_fill_discrete(name = '') +
+    labs(x = 'Hour: {0,4,8,12,16,20,23,25,39,49,63,73,97,120,144,168,192,216,240,312}', y = 'OTU richness') +
+    guides(fill=guide_legend(nrow=4,byrow=TRUE, position = 'bottom')) +
+    theme_bw() +
+    theme(
+      panel.grid.minor = element_blank(),
+      panel.grid.major = element_blank(),
+      panel.background = element_blank(),
+      axis.text.x=element_blank(),
+      legend.position = 'bottom')
+
+plot_it(p, name = 'figOTURichByCageStatus', dir = image_dir, width = 6.5, height = 5)
+
+p
+
+```
+
+(ref:figOTURichByCageStatus) Figure `r refs$figOTURichByCageStatus`
+
+`r pagebreak` ![](images\\figOTURichByCageStatus.png)
+
+**(ref:figOTURichByCageStatus):** OTU richness from communities rarefied to 5000 seqeunces, colored by OTU status.
+This figure is to compare mice in cohousing environments during the recovery process, with mice caged alone during the recovery process.
 
@@
   filter(otu %in% j$otu) %>%
-  mutate(label = paste(Family, 'OTU', as.numeric(as.factor(otu)))) %>%
+  ungroup() %>%
+  mutate(
+    label = paste(Family, 'OTU', as.numeric(as.factor(otu))),
+    Abx = ifelse(Abx == 'N', 'No abx', 'Abx')) %>%
   left_join(j) %>%
@@
   scale_y_log10() +
-  scale_color_manual(values = c('black','red')) +
+  scale_color_manual(values = c('red','black'), name = '') +
   facet_wrap(~label) + 
@@
 
+p
 ```
@@
 
-`r pagebreak` ![](images\\figImmAbunTime.png)
+`r pagebreak` ![](images\\figMostAffectedOTUs.png)
 
-**(ref:figMostAffectedOTUs):** Figure legend.
+**(ref:figMostAffectedOTUs):** Trends in abundances of 12 taxa that differ significantly between antibiotic treated mice and control mice.
+The translucent red square shows the window of antibiotic treatment.
 

### 2018-07-06 Manuscript_Mouse_Microbiome.Rmd
fixes

@@
 ```{r, include=FALSE}
+
+knitr::opts_chunk$set(cache = TRUE)
 
@@
 # load packages and custom functions
-wd <- 'C:\\Users\\John\\Documents\\msu\\mouse_microbiome\\'
+wd <- 'C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\'
+setwd(wd)
+
 source(paste0(wd, 'custom_functions.R'))
@@
                         'Present throughout experiment',
-                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
+                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144)')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
          Pair = paste('Pair', Pair),
@@
     geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
-    facet_grid(Abx~Pair) +
+    facet_grid(Pair~Abx) +
     scale_fill_manual(name = '', values = c('#a6611a','#dfc27d','#80cdc1','#018571')) +
@@
 
-plot_it(p, name = 'figOTUAbunByAbxStatus', dir = image_dir, width = 6.5, height = 5)
+plot_it(p, name = 'figOTUAbunByAbxStatus', dir = image_dir, width = 7, height = 7)
 
@@
 
-x %>%
+p <- x %>%
   ungroup() %>%
@@
                         'Present throughout experiment',
-                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
+                        'Extinct during end of Abx treatment (between Hours 63-120); returns during recovery period (after Hour 144)')[match(as.numeric(factor(status_recovery, levels = sort(unique(status_recovery)))), c(1:4))],
          Mouse = paste(ifelse(Cohoused == 'Y', 'Cohoused', 'Not cohoused'), "- Mouse", Mouse),
@@
     geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
-    facet_wrap(~Mouse, ncol = 5) +
+    facet_wrap(~Mouse, ncol = 2) +
     scale_fill_manual(name = '', values = c('#a6611a','#dfc27d','#80cdc1','#018571')) +
@@
 
-plot_it(p, name = 'figOTUAbunByCageStatus', dir = image_dir, width = 6.5, height = 5)
+plot_it(p, name = 'figOTUAbunByCageStatus', dir = image_dir, width = 7, height = 7)
 
@@
 
-plot_it(p, name = 'figOTURichByAbxStatus', dir = image_dir, width = 6.5, height = 5)
+plot_it(p, name = 'figOTURichByAbxStatus', dir = image_dir, width = 7, height = 7)
 
@@
 
-plot_it(p, name = 'figOTURichByCageStatus', dir = image_dir, width = 6.5, height = 5)
+plot_it(p, name = 'figOTURichByCageStatus', dir = image_dir, width = 7, height = 7)
 

### 2018-07-11 Manuscript_Mouse_Microbiome.Rmd
update

@@
 
-```{r, include=FALSE}
+```{r, include=FALSE, cache = FALSE}
 
-knitr::opts_chunk$set(cache = TRUE)
+knitr::opts_chunk$set(cache = FALSE)
 
@@
 
-**(ref:figOTUFates):** The potential fates and fortunes of OTUs in mice treated with antibiotics. Positive outcomes are shaded light green, whereas negative outcomes are shaded light red.
+**(ref:figOTUFates):** The potential fates and fortunes of OTUs in mice treated with antibiotics. Positive or neutral outcomes are shaded gray, whereas negative outcomes are shaded red.
 
@@
 
+if(FALSE) {
+  saveRDS(tax, file = 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\Reese2018_OTUs_wide.RDS')
+}
+
 # Note that, oddly, x[130,4719] is 'NA'. I'm assuming it is zero...?
@@
 x_wide <- x
+
+if (FALSE) {
+  saveRDS(x_wide, file = 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\Reese2018_OTUs_wide.RDS')
+}
 
@@
   expand_limits(y = c(0,1)) +
-  labs(x = 'Taxonomic level', y = 'Percent of all Phyla present across all mice')
+  labs(x = 'Taxonomic level', y = 'Percent of all Phyla present across all mice') +
   th
@@
 This hypothesis is supported.
-However, we can't be sure if the pattern is because the taxa that are more abundant taxa in the control mouse are simply better suited for the gut environment, and thus are faster to colonize the antibiotic-treated mouse.
+However, we can't be sure if the hypothesis is true because the taxa that are more abundant taxa in the control mouse may simply be better suited for the gut environment, and thus are faster to colonize the antibiotic-treated mouse.
 So, the analysis has it's limitations, but I still think it's worth noting.
@@
          Pair = paste('Pair', Pair),
-         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx')) %>%
-  ggplot(aes(x = as.factor(Hour), y = abun, fill = OTU_status)) +
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'))
+
+p <- ggplot(p, aes(x = as.factor(Hour), y = abun, fill = OTU_status)) +
     geom_bar(stat='identity', alpha = 0.8) +
-    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -Inf, ymax = 0, fill = 'red', data = filter(p, Abx == 'Abx')) +
     facet_grid(Pair~Abx) +
@@
 
-plot_it(p, name = 'figOTUAbunByAbxStatus', dir = image_dir, width = 7, height = 7)
+plot_it(p, name = 'figOTUAbunByAbxStatus', dir = image_dir, width = 7, height = 8)
 
@@
     geom_bar(stat='identity', alpha = 0.8) +
-    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -Inf, ymax = 0, fill = 'red') +
     facet_wrap(~Mouse, ncol = 2) +
@@
 
-plot_it(p, name = 'figOTUAbunByCageStatus', dir = image_dir, width = 7, height = 7)
+plot_it(p, name = 'figOTUAbunByCageStatus', dir = image_dir, width = 7, height = 8)
 
@@
          Pair = paste('Pair', Pair),
-         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx')) %>%
-  ggplot(aes(x = as.factor(Hour), fill = OTU_status)) +
+         Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'))
+
+p <- ggplot(p, aes(x = as.factor(Hour), fill = OTU_status)) +
     geom_bar(alpha = 0.8) +
-    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -Inf, ymax = 0, fill = 'red', alpha = 0.2, data = filter(p, Abx == 'Abx')) +
     facet_grid(Pair~Abx) +
@@
 
-plot_it(p, name = 'figOTURichByAbxStatus', dir = image_dir, width = 7, height = 7)
+plot_it(p, name = 'figOTURichByAbxStatus', dir = image_dir, width = 7, height = 8)
 
@@
 
-**(ref:figOTURichByAbxStatus):** OTU richness from communities rarefied to 5000 seqeunces, colored by OTU status.
-This figure is to compare mice in cohousing environments during the recovery process, with mice caged alone during the recovery process.
+**(ref:figOTURichByAbxStatus):** OTU richness from communities rarefied to 5000 seqeunces, colored by OTU status. This figure is to compare mice treated by antibiotics to untreated mice.
 
@@
     geom_bar(alpha = 0.8) +
-    geom_rect(xmin = 0, xmax = 14, ymin = -100, ymax = 0, fill = 'red', alpha = 0.2) +
+    geom_rect(xmin = 0, xmax = 14, ymin = -Inf, ymax = 0, fill = 'red') +
     facet_wrap(~Mouse, ncol = 2) +
@@
 
-plot_it(p, name = 'figOTURichByCageStatus', dir = image_dir, width = 7, height = 7)
+plot_it(p, name = 'figOTURichByCageStatus', dir = image_dir, width = 7, height = 8)
 
@@
 **(ref:figMostAffectedOTUs):** Trends in abundances of 12 taxa that differ significantly between antibiotic treated mice and control mice.
-The translucent red square shows the window of antibiotic treatment.
+The taxa were selected because they exhibited the greatest differences in abundances between treatments.
+The translucent red square highlights the window of antibiotic treatment.
 

### 2018-07-11 custom_functions.R

@@ -1,5 +1,87 @@
 # Custom functions for microbiome analyses. Created Feb 28, 2017.
 
+bac_extract <- function(x, trait) {
+  
+  # for each species query result, filter blanks and pick best value
+  for(i in 1:length(tax_bacdat)) {
+    names(x[[i]]) <- tax_bacdat[[i]]
+    x[[i]] <- x[[i]][sapply(x[[i]], length) > 0]
+    x[[i]] <- sapply(x[[i]], pick_best)
+  }
+  
+  # unlist, clean, return df
+  x <- unlist(x)
+  x[x == 'NULL'] <- NA
+  x_names <- strsplit(names(x), '.', fixed = TRUE)
+  x <- data.frame(
+    Genus = sapply(x_names, function(y) y[[1]]),
+    Species = sapply(x_names, function(y) y[[2]]),
+    trait = trait,
+    val = x
+  )
+  x <- x[!is.na(x$val), ]
+  x$val <- as.character(x$val)
+  return(x)
+}
+
+bac_search <- function (x) {
+  # needs list of genera to search for
+  
+  #just use the first page of results
+  page <- '1'
+  taxon <- 'https://bacdive.dsmz.de/api/bacdive/taxon/'
+  
+  for (i in 1:length(x)) {
+    
+    url_genus <- URLencode(paste0(taxon, x[i], '/?page=', page, '&format=json'))
+    response <- getURL(url_genus, userpwd="guittarj@msu.edu:twkYkcJbxCQzEvwkUi", httpauth = 1L)
+    jsondata <- fromJSON(response)
+    urls <- unlist(jsondata$results)
+    
+    if (length(urls) > 0) {
+      genus <- x[i] 
+      bacdat[[genus]] <- list()
+      for (j in 1:length(urls)) {
+        urlx <- URLencode(paste0(urls[j],'?&format=json'))
+        response <- getURL(urlx, userpwd="guittarj@msu.edu:twkYkcJbxCQzEvwkUi", httpauth = 1L)
+        jsondata <- fromJSON(response)
+        bacdat[[genus]][[j]] <- jsondata
+      }
+    }
+    print(paste("Downloaded", length(urls), "entries for", x[i]))
+  }
+}
+
+
+bac_search_ID <- function (x, bacdat = list()) {
+  
+  #just use the first page of results
+  page <- '1'
+  bacURL <- 'https://bacdive.dsmz.de/api/bacdive/'
+  
+  for (i in x) {
+    
[…]

### 2018-07-11 mouse_microbiome_processing_trait_data.R

@@ -1,0 +1,727 @@
+#Microbial data scubbing and organization... work in progress
+#John Guittar
+
+#set working directory, load packages
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+source('C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\custom_functions.R')
+library(data.table)
+library(tidyverse)
+
+#load full picrust/greengenes taxonomy
+ggtax <- readRDS(paste0(wd, 'ggtax.RDS'))
+ggtax <- mutate_all(ggtax, as.character)
+
+#load our OTU data; 
+#determine our taxonomy data (to know which species to focus on while webcrawling, etc)
+otus_wide <- readRDS(paste0(wd, 'Reese2018_OTUs_wide.RDS'))
+tax <- readRDS(paste0(wd, 'Reese2018_OTUs_wide.RDS'))
+
+####First: Edits drawn from from Barberan et al 2016 script
+####Then: my own edits
+
+#read table
+ijsem<-read.delim(paste0(wd, 'IJSEM_pheno_db_v1.0.txt'), sep="\t", header=T, check.names=F, fill=T,
+                  na.strings=c("NA", "", "Not indicated", " Not indicated","not indicated", "Not Indicated", "n/a", "N/A", "Na", "Not given", "not given","Not given for yeasts", "not indicated, available in the online version", "Not indicated for yeasts", "Not Stated", "Not described for yeasts", "Not determined", "Not determined for yeasts"))
+
+#simplify column names
+colnames(ijsem)<-c("Habitat", "Year", "DOI", "rRNA16S", "GC", "Oxygen",
+                   "Length", "Width", "Motility", "Spore", "MetabAssays", "Genus", "Species", "Strain", "pH_optimum", "pH_range", "Temp_optimum", "Temp_range", "Salt_optimum", "Salt_range", "Pigment", "Shape", "Aggregation", "FirstPage", "CultureCollection", "CarbonSubstrate", "Genome", "Gram", "Subhabitat", "Biolog")
+
+#clean Habitat column
+levels(ijsem$Habitat)[levels(ijsem$Habitat)=="freshwater (river, lake, pond)"]<-"freshwater"
[…]

### 2018-07-12 Manuscript_Mouse_Microbiome.Rmd
adding traits

@@
 wd <- 'C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\'
+wd_dat <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
 setwd(wd)
@@
 # load data
-x_raw <- fread('C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\2018_06_12_run2_otus.csv')
-tax <- fread('C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\otus_table_joined_taxa.txt')
+x_raw <- fread(paste0(wd_dat, '2018_06_12_run2_otus.csv'))
+tax <- fread(paste0(wd_dat, 'otus_table_joined_taxa.txt'))
+traits <- readRDS(paste0(wd_dat, 'traits_castor_subtree_averaging.RDS')) %>%
+  mutate(otu = as.integer(gsub('otu', '', otu)))
 
@@
   mutate(otu = as.numeric(otu))
+
+#filter so trait table has only the same taxa.
+#create wide version
+traits <- filter(traits, otu %in% x$otu)
+traits_wide <- spread(traits, trait, val, fill = NA)
 
@@
 
+```{r, include = FALSE}
+
+iwant <- c('Copies_16S','Forms_biofilms','Gene_number','Obligate_anaerobe','Sporulation')
+
+tmp <- x %>%
+  left_join(traits_wide[ ,c('otu', iwant)], by = 'otu') %>%
+  gather(trait, val, .dots = iwant) %>%
+  group_by(Mouse, Abx, Cohoused, Hour, trait) %>%
+  filter(Cohoused != 'dead') %>%
+  summarise(CWM = weighted.mean(val, w = abun)) %>%
+  ungroup() %>%
+  mutate(Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'),
+         Cohoused = ifelse(Cohoused == 'Y', 'Cohoused', 'Solitary'))
+
+ggplot(tmp, aes(x = Hour, y = CWM, color = Abx)) +
+    geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data =  group_by(tmp, trait, Abx) %>% filter(row_number() == 1 & Abx == 'Abx')) +
+    stat_smooth(aes(group = Abx), se = FALSE, size = 1.5) +
+    stat_smooth(aes(lty = Cohoused, group = Mouse), geom="line", se = FALSE, alpha = 0.6) +
+    scale_color_manual(values = c('red','black'), name = '') +
+    scale_linetype_discrete(name = '') +
+    facet_grid(trait~Abx, scales = 'free') +
+    guides(lty=guide_legend(override.aes = list(color = 'black'))) +
+    th
+
+```
+
 `r pagebreak` Works Cited

### 2018-07-12 custom_functions.R

@@ -24,7 +24,7 @@ bac_extract <- function(x, trait) {
   return(x)
 }
 
-bac_search <- function (x) {
+bac_search <- function (x, bacdat = list()) {
   # needs list of genera to search for
   
   #just use the first page of results
@@ -50,6 +50,7 @@ bac_search <- function (x) {
     }
     print(paste("Downloaded", length(urls), "entries for", x[i]))
   }
+  return(bacdat)
 }
 
 
@@ -65,21 +66,49 @@ bac_search_ID <- function (x, bacdat = list()) {
     response <- getURL(url_bacID, userpwd="guittarj@msu.edu:twkYkcJbxCQzEvwkUi", httpauth = 1L)
     bacdat[[i]] <- fromJSON(response)
     
-    prog <- match(i, x)
+    #Print status
+    print(paste("Downloaded", match(i, x), "of", length(x), "entries"))
+    
     if(match(i, x) %% 500 == 0) {
-      
-      #Print status
-      print(paste("Downloaded", match(i, x), "of", length(x), "entries"))
-      
+    
       #save progress
       saveRDS(bacdat, file = paste0('bacdat', match(i, x), '.RDS'))
+      
     }
+  }
+  return(bacdat)
+}
+
+pick_best <- function(x) {
   
+  #to ensure the same random sampling.
+  set.seed(7)
+  
+  # is it numeric?
+  if (is.numeric(x) & sum(!is.na(x)) > 0) {
+    
+    val <- mean(x, na.rm = T)
+    
+    # if not...
+  } else if (!is.numeric(x) & sum(!is.na(x)) > 0) {
+    
+    x <- table(as.character(x))
+    
+    # if values are of equal abundance, random sample
+    if (length(unique(x)) == 1) val <- sample(names(x), 1)
+    # otherwise select maximum value
+    if (length(unique(x)) > 1) val <- names(x)[which.max(x)]
+    
+  } else {
+    
+    val <- NA
+    
   }
   
-  return(bacdat)
-
+  return(val)
+  
 }
+
 
 
 loadpax <- function(pkg){

### 2018-07-12 mouse_microbiome_infer_missing_trait_data.R

@@ -1,0 +1,193 @@
+# In this script, I use two different methods to estimate missing trait data.
+# In the first, I use genus-level means if there are any species data available,
+# In the second, I use PICRUSt-based inference
+
+library(data.table)
+library(tidyverse)
+
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+x <- readRDS('traits_sparse.RDS')
+tax <- readRDS('ggtax.RDS') 
+
+
+##############################################################
+##################  METHOD 1: GENUS-LEVEL MEANS   #############
+##############################################################
+
+
+#calculate genus-level means
+x_genus_means <- x %>%
+  gather(trait, val, -otu) %>%
+  left_join(tax[, c('Genus','otu')], by = c('otu')) %>%
+  group_by(trait, Genus) %>%
+  filter(sum(!is.na(val)) > 0) %>%
+  mutate(val = ifelse(is.na(val), mean(val, na.rm = T), val)) %>%
+  spread(trait, val) %>%
+  ungroup() %>%
+  select(-Genus)
+
+saveRDS(x_genus_means, file = 'traits_genus_means.RDS')
+
+
+
+
+
+##############################################################
+##################  METHOD 2: PICRUSt   ######################
+##############################################################
+
+
+# Finally figured out how to use the picrust algorithm to estimate unknown trait values
+# I manually ran dataframes with known trait values through picrust on the hpcc. First, create the trait input files
+#
+if (FALSE) {
+  x_tmp <- x %>%
+    gather(trait, val, -otu) %>%
+    filter(!is.na(val))
+  
+  for (i in unique(x_tmp$trait)) {
+    tmp <- x_tmp %>%
+      filter(trait == i) %>%
+      mutate(
+        otu = as.numeric(gsub("otu", "", otu)), 
+        trait = NULL)
+    colnames(tmp) <- c('GenomeID', i)
+    write.table(tmp, file = paste0('data\\picrust_input_traits\\', i, '.tab'), row.names = FALSE, sep = '\t', quote = FALSE)
+  }
+}
+
+##then, on the HPCC:
+## open hpcc terminal
[…]

### 2018-07-12 mouse_microbiome_processing_trait_data.R

@@ -9,13 +9,13 @@ library(data.table)
 library(tidyverse)
 
 #load full picrust/greengenes taxonomy
-ggtax <- readRDS(paste0(wd, 'ggtax.RDS'))
+ggtax <- readRDS('ggtax.RDS')
 ggtax <- mutate_all(ggtax, as.character)
 
 #load our OTU data; 
 #determine our taxonomy data (to know which species to focus on while webcrawling, etc)
-otus_wide <- readRDS(paste0(wd, 'Reese2018_OTUs_wide.RDS'))
-tax <- readRDS(paste0(wd, 'Reese2018_OTUs_wide.RDS'))
+otus_wide <- readRDS('Reese2018_OTUs_wide.RDS')
+tax <- readRDS('Reese2018_OTUs_wide.RDS')
 
 ####First: Edits drawn from from Barberan et al 2016 script
 ####Then: my own edits
@@ -207,7 +207,7 @@ ijsem <- select(ijsem, Genus, Species, GC_content = GC,
 ###########################################################
 # now, sporulation data from Browne et al 2016
 
-spo <- read.csv(paste0(wd, 'Browne2016_sporulationTable.csv'), header = T, skip = 1)
+spo <- read.csv('Browne2016_sporulationTable.csv', header = T, skip = 1)
 
 # fix long colnames
 colnames(spo) <- substr(colnames(spo), 1, 2)
@@ -221,14 +221,14 @@ spo <- spo %>%
 
 ####################################### Exploring BacDrive
 
-library(RCurl)
-library(rjson)
-
 if (FALSE) {
+  #IFF i want just the genera in the mouse experiment:
+  library(RCurl)
+  library(rjson)
   
   #needed <- sort(as.character(unique(filter(tax, is.na(Gram_positive))$Genus)))
   needed <- sort(unique(tax$Genus))
-  needed <- needed[!needed %in% c("unclassified","SMB53","02d06","rc4-4","WAL_1855D","1-68","cc_115")]
+  needed <- needed[!needed %in% c("unclassified","SMB53","02d06","rc4-4","WAL_1855D","1-68","cc_115","KD1-23")]
   
   bacdat <- bac_search(needed) #re-calculate bacdata
   saveRDS(bacdat, file = paste0(wd, 'bacdive_genera_', Sys.Date(), '.RDS'))
@@ -237,37 +237,50 @@ if (FALSE) {
[…]

### 2018-07-13 Manuscript_Mouse_Microbiome.Rmd
refining traits

@@
   summarise(abun = sum(abun)) %>%
-  mutate(otu = as.numeric(otu))
+  mutate(otu = as.numeric(otu)) %>%
+  group_by(otu) %>%
+  filter(sum(abun) > 0) %>%
+  ungroup()
 
-#filter so trait table has only the same taxa.
+#filter so trait and tax tables have the same taxa.
 #create wide version
@@
 traits_wide <- spread(traits, trait, val, fill = NA)
+tax <- filter(tax, otu %in% x$otu)
 
@@
     panel.background = element_blank())
+
+#set trait names for plotting
+#renaming
+trait_names <- c(
+  "B_vitamins" = "B vitamins",
+  "Copies_16S" = "16S gene copies",
+  "Forms_biofilms" = "Forms biofilms",
+  "GC_content" = "GC content",
+  "Gene_number" = "Genes",
+  "Gram_positive" = "Gram-positive",
+  "IgA" = "IgA binding affinity",
+  "Obligate_anaerobe" = "Obligate anaerobe",
+  "Temp_optimum" = "Temperature optimum"
+)
 
@@
 
-ggplot(tmp, aes(x = Hour, y = CWM, color = Abx)) +
+p <- ggplot(tmp, aes(x = Hour, y = CWM, color = Abx)) +
     geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data =  group_by(tmp, trait, Abx) %>% filter(row_number() == 1 & Abx == 'Abx')) +
@@
     facet_grid(trait~Abx, scales = 'free') +
-    guides(lty=guide_legend(override.aes = list(color = 'black'))) +
+    guides(lty=guide_legend(override.aes = list(color = 'black', alpha = 1))) +
     th
+
+plot_it(p, name = 'figFiveTraitPatterns', dir = image_dir, width = 6, height = 7.5)
+
+p
 
@@
 
+(ref:figFiveTraitPatterns) Figure `r refs$figFiveTraitPatterns`
+
+`r pagebreak` ![](images\\figFiveTraitPatterns.png)
+
+**(ref:figFiveTraitPatterns):** Shifts in community weighted means of mouse gut microbial communities during and after antibiotic treatment (left) or in untreated control mice (right).
+Thick colored lines show average CWM trait values across mice, whereas thin colored lines show CWM trends for individual mice.
+
+```{r, include = FALSE}
+
+ggtax <- readRDS(paste0(wd_dat, 'ggtax.RDS')) %>%
+  mutate(otu = as.integer(gsub('otu', '', otu)))
+
+tmp <- traits_wide %>%
+  gather(trait, val, -otu) %>%
+  filter(!is.na(val)) %>%
+  left_join(ggtax, by = 'otu') %>%
+  filter(Species != 'unclassified')
+
+tmp_abuns <- x %>% ungroup() %>% group_by(otu) %>% summarise(abun = sum(abun))
+
+#I drop traits that are already dependent on inferred trait data, from Bugbase
+j <- traits_wide %>%
+  select(-Forms_biofilms, -Obligate_anaerobe, -Gram_positive) %>%
+  gather(trait, val, -otu) %>%
+  distinct(otu, trait) %>%
+  inner_join(tax, by = 'otu') %>%
+  mutate(
+    Species = paste(trait, Family, Genus, Species) %in% paste(tmp$trait, tmp$Family, tmp$Genus, tmp$Species),
+    Genus = paste(trait, Family, Genus) %in% paste(tmp$trait, tmp$Family, tmp$Genus),
+    Family = paste(trait, Family) %in% paste(tmp$trait, tmp$Family),
+    Order = paste(trait, Order) %in% paste(tmp$trait, tmp$Order),
+    Class = paste(trait, Class) %in% paste(tmp$trait, tmp$Class)
+  ) %>%
+  rename(C = Class, O = Order, F = Family, G = Genus, S = Species) %>%
+  gather(level, coverage, C, O, F, G, S) %>%
+  inner_join(tmp_abuns, by = 'otu') %>%
+  mutate(level = factor(level, levels = rev(c('C','O','F','G','S')))) %>%
+  group_by(trait, level)
+
+# individuals with trait data by trait and level
+#j %>% 
+#  group_by(trait, level) %>% 
+#  summarise(abun = sum(abun[coverage])) %>%
+#  spread(level, abun)
+
+#percent individuals with trait data by trait and level
+#j %>% 
+#  group_by(trait, level) %>% 
+#  summarise(abun = round(sum(abun[coverage]) / sum(abun), 2)) %>%
+#  spread(level, abun)
+
+# summary across traits
+#j %>% 
+#  group_by(level) %>% 
+#  summarise(abun = sum(abun[coverage])) %>%
+#  mutate(abun_percent = abun / max(abun))
+
+# numbers of OTUs with trait data by trait and level
+#j %>% 
+#  group_by(trait, level) %>% 
+#  summarise(rich = sum(coverage)) %>%
+#  spread(level, rich)
+
+# numbers of OTUs with trait data by level
+#j %>% 
+#  group_by(trait, level) %>% 
+#  summarise(rich = round(sum(coverage) / length(coverage), 2)) %>%
+#  spread(level, rich)
+
+# summary across traits
+#j %>% 
+#  group_by(level) %>% 
+#  summarise(rich = sum(coverage)) %>%
+#  mutate(rich_percent = rich / max(rich))
+
+p <- j %>%
+  summarise(coverage = sum(abun[coverage]) / sum(abun)) %>%
+  filter(trait %in% iwant) %>%
+  ungroup() %>%
+  mutate(trait = ifelse(trait %in% names(trait_names), trait_names[match(trait, names(trait_names))], trait)) %>%
+  ggplot(aes(x = level, y = coverage)) +
+    geom_bar(stat = 'identity') +
+    facet_wrap(~trait) +
+    theme_bw() +
+    theme(
+      panel.grid.minor = element_blank(),
+      panel.grid.major = element_blank(),
+      panel.background = element_blank(),
+      legend.position = 'none') +
+    labs(x = 'Maximum level of phylogenetic distance allowed when inferring trait data', y = 'Proportion of community with available trait data')
+
+plot_it(p, name = 'figCoverage', dir = image_dir, width = 6, height = 3.5)
+
+p
+
+```
+
+(ref:figCoverage) Figure `r refs$figCoverage`
+
+`r pagebreak` ![](images\\figCoverage.png)
+
+**(ref:figCoverage):** 
+Unknown trait data were inferred by averaging trait values over descending subtrees in the greengenes phylogeny.
+That is, for each unknown trait value, the predicted value was set to the arithmetic average state of all tips descending from that node.
+On the x-axis, S = Species, G = Genus, F = Family, O = Order, C = Class.
+Missing trait data for three traits (the functional potential to form biofilms, obgligate anaerobe status, and Gram-positive status) were inferred for all OTUs in the greengenes phylogeny in Ward et al.
+2017, and are therefore not reported in this figure.
+
 `r pagebreak` Works Cited

### 2018-07-16 Manuscript_Mouse_Microbiome.Rmd
update

@@
 1. Which traits (other than antibiotic resistance) predict OTU tolerance/susceptibility to antibiotics?
+ + Can they explain differential microbiome community responses to antibiotics?
 
@@
 
+tmp <- select(x_wide, -Mouse, -Abx, -Dose, -Hour, -Cohoused, -Pair)
+
 #####
-#takes too long
-#ord <- metaMDS(x_wide, k = 3, trymax = 1000, noshare = 0.2)
+#doesn't easily converge... but based on a quick test the results don't differ too much.
+ord <- metaMDS(tmp)
+ord_scores <- as.data.frame(scores(ord))
 #####
 
-myp <- select(x_wide, -Mouse, -Abx, -Dose, -Hour, -Cohoused, -Pair)
-myp <- pcoa(vegdist(myp))
+myp <- pcoa(vegdist(tmp))
+# to see how much of
 
@@
   PCoA1 = myp$vectors[,1],
-  PCoA2 = myp$vectors[,2]) %>%
+  PCoA2 = myp$vectors[,2],
+  NMDS1 = ord_scores$NMDS1,
+  NMDS2 = ord_scores$NMDS2) %>%
   group_by(Mouse) %>%
@@
   ggplot(aes(x = PCoA1, y = PCoA2, color = Abx, fill = Abx)) + 
+  geom_point(aes(alpha = Hour), shape = 21, size = 4) + 
+  geom_point(shape = 21, size = 4, fill = NA) + 
+  geom_path() + 
+  geom_text(aes(label = label), color = 'black', na.rm=TRUE) +
+  scale_color_manual(values = c('black','red')) +
+  scale_fill_manual(values = c('black','red')) +
+  facet_grid(Pair~Abx) + 
+  th
+
+p2 <- j %>% 
+  filter(!is.na(Pair)) %>%
+  mutate(Pair = paste('Cage', Pair)) %>%
+  ggplot(aes(x = NMDS1, y = NMDS2, color = Abx, fill = Abx)) + 
   geom_point(aes(alpha = Hour), shape = 21, size = 4) + 
@@
 
-iwant <- c('Copies_16S','Forms_biofilms','Gene_number','Obligate_anaerobe','Sporulation')
+iwant <- c('Obligate_anaerobe','Sporulation')
 
@@
 
-plot_it(p, name = 'figFiveTraitPatterns', dir = image_dir, width = 6, height = 7.5)
+plot_it(p, name = 'figFiveTraitPatterns', dir = image_dir, width = 6, height = 6)
 

### 2018-09-06 Manuscript_Mouse_Microbiome.Rmd
major update post review

@@
 
-John Guittar^1,\*^, Aspen Reese^2^, Lawrence David^3^, Timothy Brennhofer^1,4^, Ashley Shade^5,6^, Elena Litchman^1,6,7^
+John Guittar^1,\*^, Aspen Reese^2^, Lawrence David^3^, Ashley Shade^5,6^, Elena Litchman^1,6,7^
 
@@
 
-^4^ Grinnell College, 1115 8th Ave, Grinnell, IA 50112, USA
-
-^5^ Department of Microbiology and Molecular Genetics, Department of Plant, Soil and Microbial Sciences, and The Plant Resilience Institute, Michigan State University, East Lansing MI 48840, USA
-
-^6^ Program in Ecology, Evolutionary Biology and Behavior, Michigan State University East Lansing MI 48840, USA
-
-^7^ Department of Integrative Biology, Michigan State University, East Lansing, MI 48824, USA
+^4^ Department of Microbiology and Molecular Genetics, Department of Plant, Soil and Microbial Sciences, and The Plant Resilience Institute, Michigan State University, East Lansing MI 48840, USA
+
+^5^ Program in Ecology, Evolutionary Biology and Behavior, Michigan State University East Lansing MI 48840, USA
+
+^6^ Department of Integrative Biology, Michigan State University, East Lansing, MI 48824, USA
 
@@
 #rarefy to 5000
+if (FALSE) {
   tmp <- x[,c(11:ncol(x))]
@@
   x <- bind_cols(x[, c(1:10)], as.data.frame(tmp))
+}
 
@@
 traits_wide <- spread(traits, trait, val, fill = NA)
+traits_wide_scaled <- data.frame(otu = traits_wide$otu, sapply(traits_wide[, -1], scale))
+traits_wide_scaled <- select(traits_wide_scaled, -Length, -B_vitamins, -Temp_optimum)
 tax <- filter(tax, otu %in% x$otu)
@@
   scale_fill_manual(values = c('black','red')) +
-  facet_grid(Pair~Abx) + 
+  facet_wrap(~Mouse) + 
   th
@@
   group_by(Mouse, Hour) %>%
-  mutate(rich = length(unique(otu[abun > 0]))) %>%
-  group_by(Mouse, Hour, Abx, Cohoused, Dose, Pair, rich) %>%
+  left_join(tax[, c('otu','Genus','Species')], by = 'otu') %>%
+  mutate(rich_spp = length(unique(paste(Genus, Species)[abun > 0 & Species != 'unclassified'])),
+         rich = length(unique(otu[abun > 0]))) %>%
+  group_by(Mouse, Hour, Abx, Cohoused, Dose, Pair, rich, rich_spp) %>%
   do(shannon = diversity(spread(select(., otu, abun), otu, abun))) %>%
@@
 
-#must fix screwy data!!! (when I paste mouse, abx, cohoused, there are some unexpected combinations)
 j <- j %>%
-  group_by(Mouse, Abx, Cohoused) %>%
-  filter(length(shannon_val) > 1) %>%
   ungroup() %>%
@@
     geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data = j %>% group_by(Mouse) %>% filter(row_number() == 1 & Abx == 'Abx')) + 
+    geom_hline(yintercept = 0) +
     facet_wrap(~Mouse, ncol = 3) +
     labs(x = 'Hour', y = 'OTU richness') +
-    th
+    theme_bw()
 
@@
 #set the window. The edges are included.
-window0 <- 63 
+window0 <- 119 
 window1 <- 120
-
-j <- x %>%
-  group_by(Mouse, Abx, Pair, otu) %>%
-  summarise(
-    abunPre = if (unique(Abx) == 'Y') sum(abun[Hour <= window0]) else NA,
-    abun0 = if (unique(Abx) == 'Y') sum(abun[Hour >= window0 & Hour <= window1]) else sum(abun[Hour < window1]),
-    abun1 = sum(abun[Hour > window1]),
-    statusAB = if (!is.na(abunPre) & abunPre == 0 & abun0 == 0) 'absent' else 'persistent',
-    statusAB = if (!is.na(abunPre) & abunPre > 0 & abun0 == 0) 'extinct' else statusAB,
-    ETD = if (statusAB == 'extinct') max(Hour[Hour <= window0 & abun > 0]) else NA,
-    ETD = if (statusAB == 'persistent') Inf else ETD,
-    status0 = if (abun0 > 0) 'present' else 'absent',
-    status1 = if (sum(abun[Hour > window1] > 0)) 'present' else 'absent',
-    status = if (status0 == 'absent' & status1 == 'absent') 'absent' else 'persistent',
-    status = if (status0 == 'absent' & status1 == 'present') 'immigrant' else status,
-    status = if (status0 == 'present' & status1 == 'absent') 'extinct' else status,
-    ETA = if (status == 'immigrant') min(Hour[Hour > 97 & abun > 0]) else NA) %>%
-  group_by(Pair, otu) %>%
-  mutate(
-    abun_pair = ifelse(!is.na(Pair) & Abx == 'Y', abun0[Abx == 'N'], NA),
-    abun_pair = ifelse(!is.na(Pair) & Abx == 'N', abun0[Abx == 'Y'], abun_pair))
 
@@
     abun2 = sum(abun[Hour > window1]),
+    abun_final = if (312 %in% Hour) abun[Hour == 312]
+                   else NA,
     status_treatment = if (abun01 == 0) 'absent' 
@@
 p <- j %>% 
-  filter(status_recovery == 'immigrant' & Abx == 'Y' & abun_pair > 0) %>%
+  filter(status_recovery == 'immigrant' & Abx == 'Y') %>%
   mutate(time_to_arrival = ETA - 120) %>%
-  ggplot(aes(x = abun_pair, y = time_to_arrival)) + 
+  ggplot(aes(x = abun_pair + 1, y = time_to_arrival)) + 
     geom_jitter(alpha = 0.25, color = 'red', height = 5, width = 0) + 
@@
     stat_smooth(color = 'black', lty = 3) +
-    labs(x = "OTU relative abundance in cohoused control mouse", y = 'Hours before OTU arrival in Abx treated mouse') +
+    labs(x = "OTU relative abundance in cohoused control mouse + 1", y = 'Hours before OTU arrival in Abx treated mouse') +
     th
@@
 
-iwant <- c('Obligate_anaerobe','Sporulation')
+iwant <- c('Obligate_anaerobe','Sporulation', 'Copies_16S', 'Gene_number','Motility')
 
@@
 
+```{r, include = FALSE}
+
+p <- traits %>% 
+  left_join(tax, by = 'otu') %>% 
+  group_by(trait) %>% 
+  arrange(val) %>% 
+  mutate(
+    rank = seq_along(val), 
+    Enterobacteriaceae = Family == 'Enterobacteriaceae') %>% 
+  ggplot(aes(x = rank, y = val, color = Enterobacteriaceae)) + 
+    geom_point(aes(size = Enterobacteriaceae)) + 
+    facet_wrap(~trait, scales = 'free') + 
+    scale_color_manual(values = c('black','red')) + 
+    th
+
+plot_it(p, name = 'figEnteroTraits', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figEnteroTraits) Figure `r refs$figEnteroTraits`
+
+`r pagebreak` ![](images\\figEnteroTraits.png)
+
+**(ref:figEnteroTraits):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  group_by(Mouse) %>%
+  filter(abun > 0) %>%
+  filter(Cohoused != 'dead') %>%
+  group_by(Mouse, Hour) %>%
+  mutate(rich = length(unique(otu))) %>%
+  group_by(Mouse) %>%
+  mutate(
+    cum_rich = as.numeric(factor(otu, levels = unique(otu))),
+    Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'),
+    Cohoused = ifelse(Cohoused == 'Y', 'Cohoused', 'Solitary')) %>%
+  group_by(Mouse, Hour, Abx, Cohoused, rich) %>%
+  summarise(cum_rich = max(cum_rich)) %>%
+  group_by(Mouse) %>%
+  mutate(new_rich = c(cum_rich[1], cum_rich[2:length(cum_rich)] -
+                                   cum_rich[1:length(cum_rich) -1]))
+
+p <- ggplot(j, aes(x = Hour, y = cum_rich, color = Abx)) +
+    geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data = filter(ungroup(j), row_number() == 1 & Abx == 'Abx')) + 
+    geom_vline(xintercept = 120) +
+    geom_vline(xintercept = 0) +
+    #geom_line(aes(group = Mouse, lty = Cohoused)) +
+    geom_point(data = filter(j, cum_rich < 2300)) +
+    geom_point(data = filter(j, cum_rich > 2300), shape = 1) +
+    stat_smooth(aes(group =  Mouse, lty = Cohoused), se = FALSE,
+                data = filter(j, cum_rich < 2300)) +
+    scale_color_manual(values = c('red','black')) +
+    #facet_wrap(~Abx) +
+    expand_limits(y = 0) +
+    th +
+    labs(x = "Hour", y = 'Cumulative no. of unique OTUs encountered')
+
+plot_it(p, name = 'figCumOTUS', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figCumOTUS) Figure `r refs$figCumOTUS`
+
+`r pagebreak` ![](images\\figCumOTUS.png)
+
+**(ref:figCumOTUS):**
+
+```{r, include = FALSE}
+
+p <- ggplot(j, aes(x = Hour, y = cum_rich, color = Abx, lty = Cohoused)) +
+    geom_point(alpha = 0.2, data = filter(j, cum_rich < 2300)) +
+    geom_point(data = filter(j, cum_rich > 2300)) +
+    stat_smooth(se = FALSE, data = filter(j, cum_rich < 2300)) + 
+    th +
+    scale_color_manual(values = c('red','black')) +
+    expand_limits(y = 0)
+
+plot_it(p, name = 'figCumOtusByTreat', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figCumOtusByTreat) Figure `r refs$figCumOtusByTreat`
+
+`r pagebreak` ![](images\\figCumOtusByTreat.png)
+
+**(ref:figCumOtusByTreat):**
+
+```{r, include = FALSE}
+
+p <- j %>% 
+  group_by(Mouse, Abx, Cohoused, Hour) %>%
+  summarise(rich = unique(rich)) %>%
+  ggplot(aes(x = Hour, y = rich, color = Abx, lty = Cohoused)) +
+    geom_point(alpha = 0.2) +
+    stat_smooth() + 
+    scale_color_manual(values = c('red','black')) +
+    th +
+    labs(y = 'Instantaneous richness')
+
+plot_it(p, name = 'figRichByTreat', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figRichByTreat) Figure `r refs$figRichByTreat`
+
+`r pagebreak` ![](images\\figRichByTreat.png)
+
+**(ref:figRichByTreat):**
+
+```{r, include = FALSE}
+
+p <- j %>% 
+  ggplot(aes(x = Hour, y = new_rich, color = Abx, lty = Cohoused)) +
+    geom_point(alpha = 0.2, data = filter(j, new_rich > 0 & (new_rich < 200 | Hour < 200))) +
+    geom_point(size = 2, data = filter(j, new_rich > 200 & Hour > 200)) +
+    scale_color_manual(values = c('red','black')) +
+    stat_smooth(se = FALSE, data = filter(j, new_rich < 200 | Hour < 200)) + 
+    th +
+    labs(y = 'No. new OTUs encountered')
+
+#j %>% 
+#  group_by(Mouse) %>%
+#  filter(Hour %in% c(0,23,63,120,240,312)) %>%
+#  group_by(Mouse, Abx, Cohoused, Hour) %>%
+#  summarise(cum_rich = max(cum_rich)) %>%
+#  ggplot(aes(x = factor(Hour), y = cum_rich, color = Abx)) +
+#    geom_boxplot(position = 'dodge') +
+#    facet_wrap(~Cohoused) +
+#    th
+
+plot_it(p, name = 'figNewRichByTreat', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figNewRichByTreat) Figure `r refs$figNewRichByTreat`
+
+`r pagebreak` ![](images\\figNewRichByTreat.png)
+
+**(ref:figNewRichByTreat):**
+
+```{r, include = FALSE}
+
+j <- x %>% 
+  left_join(tax, by = 'otu') %>% 
+  filter(Family == 'Enterobacteriaceae') %>%
+  group_by(Mouse) %>%
+  filter(abun > 0) %>%
+  mutate(Cumulative = cummax(as.numeric(factor(otu, levels = unique(otu))))) %>%
+  mutate(Abx = factor(ifelse(Abx == 'Y','Abx', 'No abx'), levels = c('No abx', 'Abx'))) %>%
+  group_by(Mouse, Abx, Hour) %>%  
+  summarise(
+    Instantaneous = length(otu), 
+    Cumulative = max(Cumulative)) %>%
+  gather(var, val, Instantaneous, Cumulative) %>%
+  mutate(var = factor(var, unique(var)))
+
+p <- ggplot(j, aes(x = Hour, y = val, color = Abx)) + 
+  geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', 
+            color = NA, data = filter(group_by(j, var), row_number() == 1 & Abx == 'Abx')) + 
+  geom_vline(xintercept = 0) +
+  geom_hline(yintercept = 0) +
+  #geom_line(aes(group = Mouse)) + 
+  geom_point(alpha = 0.2) +
+  stat_smooth(aes(group = Mouse), se = FALSE) +
+  scale_color_manual(values = c('black','red'), name = '') +
+  labs(y = "Enterobacteriaceae OTU richness") +
+  facet_wrap(~var) +
+  th
+
+plot_it(p, name = 'figEntRichHour', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figEntRichHour) Figure `r refs$figEntRichHour`
+
+`r pagebreak` ![](images\\figEntRichHour.png)
+
+**(ref:figEntRichHour):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  left_join(tax, by = 'otu') %>%
+  filter(Abx == 'Y') %>%
+  mutate(abun = abun/sum(abun),
+         Mouse = paste(Mouse, Cohoused)) %>%
+  group_by(Family) %>%
+  mutate(fam_abun = sum(abun)) %>%
+  ungroup() %>%
+  filter(fam_abun > 0.02) %>%
+  group_by(Mouse, Family, Hour) %>%
+  summarise(abun = sum(abun)) %>%
+  ungroup()
+
+p <- ggplot(j, aes(x = Hour, y = abun, color = Family)) +
+  geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.15, fill = 'red', color = NA, data = j %>% filter(row_number() == 1)) + 
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  #facet_wrap(~Mouse) + 
+  theme(legend.position = 'bottom',
+    panel.grid.minor = element_blank(),
+    panel.grid.major = element_blank(),
+    panel.background = element_blank())
+
+plot_it(p, name = 'figFamilyPatterns', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figFamilyPatterns) Figure `r refs$figFamilyPatterns`
+
+`r pagebreak` ![](images\\figFamilyPatterns.png)
+
+**(ref:figFamilyPatterns):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  filter(Abx == 'Y') %>%
+  group_by(Mouse, otu) %>%
+  filter(abun[Hour == 0] > 0) %>%
+  summarise(abx_response = ifelse(abun[Hour == 120] > 0, 1, 0)) %>%
+  ungroup() %>%
+  left_join(traits_wide_scaled, by = 'otu') %>%
+  select(-Mouse, -otu)
+
+m <- lm(abx_response ~ ., data = j)
+
+tmp <- summary(m)$coef %>%
+  as.data.frame() %>%
+  do(data.frame(var = row.names(.), .)) %>%
+  rename(SE = Std..Error, p.value = Pr...t..) %>%
+  filter(var != "(Intercept)") %>%
+  mutate(var = factor(var, unique(var[order(t.value)])))
+
+p <- ggplot(tmp, aes(x = var, y = Estimate, color = p.value < 0.05)) +
+  geom_hline(yintercept = 0) +
+  geom_errorbar(aes(ymin = Estimate - SE, ymax = Estimate + SE), width = 0.3) +
+  geom_point() +
+  labs(x = '', y = 'Mean change in probability of persisting (Y/N) during Abx treatment') +
+  coord_flip() +
+  th
+
+plot_it(p, name = 'figTraitsDuringAbxBinary', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figTraitsDuringAbxBinary) Figure `r refs$figTraitsDuringAbxBinary`
+
+`r pagebreak` ![](images\\figTraitsDuringAbxBinary.png)
+
+**(ref:figTraitsDuringAbxBinary):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  filter(Abx == 'Y') %>%
+  group_by(Mouse, otu) %>%
+  filter(abun[Hour == 0] > 0 | abun[Hour == 120] > 0) %>%
+  summarise(delta_rel_abun = abun[Hour == 120] - abun[Hour == 0]) %>%
+  ungroup() %>%
+  left_join(traits_wide_scaled) %>%
+  select(-Mouse, -otu)
+
+m <- lm(delta_rel_abun / 5000 ~ ., data = j)
+
+tmp <- summary(m)$coef %>%
+  as.data.frame() %>%
+  do(data.frame(var = row.names(.), .)) %>%
+  rename(SE = Std..Error, p.value = Pr...t..) %>%
+  filter(var != "(Intercept)") %>%
+  mutate(var = factor(var, unique(var[order(t.value)])))
+
+p <- ggplot(tmp, aes(x = var, y = Estimate, color = p.value < 0.05)) +
+  geom_hline(yintercept = 0) +
+  geom_errorbar(aes(ymin = Estimate - SE, ymax = Estimate + SE), width = 0.3) +
+  geom_point() +
+  labs(y = 'Mean Change in rel. abun. over Abx treatment', x = '') +
+  coord_flip() +
+  th
+
+plot_it(p, name = 'figTraitsDuringAbxContinuous', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figTraitsDuringAbxContinuous) Figure `r refs$figTraitsDuringAbxContinuous`
+
+`r pagebreak` ![](images\\figTraitsDuringAbxContinuous.png)
+
+**(ref:figTraitsDuringAbxContinuous):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  filter(!is.na(Pair)) %>%
+  group_by(Pair, otu) %>%
+  mutate(Control_abun = sum(abun[Abx == 'N']) / length(unique(Hour))) %>%
+  filter(Abx == 'Y') %>%
+  group_by(Mouse, otu, Control_abun) %>%
+  filter(abun[Hour == 120] > 0) %>%
+  summarise(abx_response = ifelse(abun[Hour == max(Hour)] > 0, 1, 0)) %>%
+  ungroup() %>%
+  left_join(traits_wide_scaled, by = 'otu') %>%
+  mutate(Control_abun = scale(Control_abun)) %>%
+  select(-Mouse, -otu)
+
+m <- lm(abx_response ~ ., data = j)
+
+tmp <- summary(m)$coef %>%
+  as.data.frame() %>%
+  do(data.frame(var = row.names(.), .)) %>%
+  rename(SE = Std..Error, p.value = Pr...t..) %>%
+  filter(var != "(Intercept)") %>%
+  mutate(var = factor(var, unique(var[order(Estimate)])))
+
+p <- ggplot(tmp, aes(x = var, y = Estimate, color = p.value < 0.05)) +
+  geom_hline(yintercept = 0) +
+  geom_errorbar(aes(ymin = Estimate - SE, ymax = Estimate + SE), width = 0.3) +
+  geom_point() +
+  labs(x = '', y = 'Mean change in probability of persisting (Y/N) over Abx recovery') +
+  coord_flip() +
+  th
+
+plot_it(p, name = 'figTraitsAfterAbxBinary', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figTraitsAfterAbxBinary) Figure `r refs$figTraitsAfterAbxBinary`
+
+`r pagebreak` ![](images\\figTraitsAfterAbxBinary.png)
+
+**(ref:figTraitsAfterAbxBinary):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  filter(!is.na(Pair)) %>%
+  group_by(Pair, otu) %>%
+  mutate(Control_abun = sum(abun[Abx == 'N']) / length(unique(Hour))) %>%
+  filter(Abx == 'Y') %>%
+  group_by(Mouse, otu, Control_abun) %>%
+  filter(abun[Hour == 120] > 0) %>%
+  summarise(delta_rel_abun = abun[Hour == 120] - abun[Hour == max(Hour)]) %>%
+  ungroup() %>%
+  left_join(traits_wide_scaled, by = 'otu') %>%
+  mutate(Control_abun = scale(Control_abun)) %>%
+  select(-Mouse, -otu)
+
+m <- lm(delta_rel_abun ~ ., data = j)
+
+tmp <- summary(m)$coef %>%
+  as.data.frame() %>%
+  do(data.frame(var = row.names(.), .)) %>%
+  rename(SE = Std..Error, p.value = Pr...t..) %>%
+  filter(var != "(Intercept)") %>%
+  mutate(var = factor(var, unique(var[order(Estimate)])))
+
+p <- ggplot(tmp, aes(x = var, y = Estimate, color = p.value < 0.05)) +
+  geom_hline(yintercept = 0) +
+  geom_errorbar(aes(ymin = Estimate - SE, ymax = Estimate + SE), width = 0.3) +
+  geom_point() +
+  labs(y = 'Mean Change in rel. abun. over Abx recovery', x = '') +
+  coord_flip() +
+  th
+
+plot_it(p, name = 'figTraitsAfterAbxContinuous', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figTraitsAfterAbxContinuous) Figure `r refs$figTraitsAfterAbxContinuous`
+
+`r pagebreak` ![](images\\figTraitsAfterAbxContinuous.png)
+
+**(ref:figTraitsAfterAbxContinuous):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  filter(!is.na(Pair)) %>%
+  group_by(Pair, otu) %>%
+  mutate(Control_abun = sum(abun[Abx == 'N']) / length(unique(Hour))) %>%
+  group_by(Mouse, otu, Control_abun) %>%
+  summarise(
+    Initial_abun = abun[Hour == 0],
+    End_abx_abun = abun[Hour == 120],
+    final_abun = abun[Hour == max(Hour)],
+    Persistence = ifelse(abun[Hour == 0] > 0 & abun[Hour == 120] > 0, 1, 0),
+    Persistence = ifelse(sum(abun[Hour %in% c(0,120)]) == 0, NA, Persistence)) %>%
+  ungroup() %>%
+  mutate(Control_abun = scale(log10(Control_abun+1)),
+         Initial_abun = scale(log10(Initial_abun+1)),
+         End_abx_bun = scale(log10(End_abx_abun+1)),
+         Persistence = scale(Persistence)) %>%
+  select(-Mouse, -otu)
+
+m <- lm(final_abun ~ ., data = j)
+
+tmp <- summary(m)$coef %>%
+  as.data.frame() %>%
+  do(data.frame(var = row.names(.), .)) %>%
+  rename(SE = Std..Error, p.value = Pr...t..) %>%
+  filter(var != "(Intercept)") %>%
+  mutate(var = factor(var, unique(var[order(Estimate)])))
+
+p <- ggplot(tmp, aes(x = var, y = Estimate, color = p.value < 0.05)) +
+  geom_hline(yintercept = 0) +
+  geom_errorbar(aes(ymin = Estimate - SE, ymax = Estimate + SE), width = 0.3) +
+  geom_point() +
+  labs(x = '', y = 'Predictors of post-recovery rel. abundance') +
+  coord_flip() +
+  th
+
+plot_it(p, name = 'figAbxPlusRecoveryPredictors', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figAbxPlusRecoveryPredictors) Figure `r refs$figAbxPlusRecoveryPredictors`
+
+`r pagebreak` ![](images\\figAbxPlusRecoveryPredictors.png)
+
+**(ref:figAbxPlusRecoveryPredictors):**
+
+```{r, include = FALSE}
+
+j <- x %>%
+  group_by(Mouse) %>%
+  filter(abun > 0) %>%
+  filter(Cohoused != 'dead') %>%
+  group_by(Mouse) %>%
+  mutate(
+    cum_rich = as.numeric(factor(otu, levels = unique(otu))),
+    Abx = ifelse(Abx == 'Y', 'Abx', 'No abx'),
+    Cohoused = ifelse(Cohoused == 'Y', 'Cohoused', 'Solitary')) %>%
+  group_by(Mouse, Hour, Abx, Cohoused) %>%
+  summarise(cum_rich = max(cum_rich))
+
+p <- ggplot(j, aes(x = Hour, y = cum_rich, color = Abx)) +
+    geom_rect(aes(xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf), alpha = 0.3, fill = 'red', color = NA, data = filter(ungroup(j), row_number() == 1 & Abx == 'Abx')) + 
+    geom_vline(xintercept = 120) +
+    geom_vline(xintercept = 0) +
+    #geom_line(aes(group = Mouse, lty = Cohoused)) +
+    geom_point(data = filter(j, cum_rich < 2300)) +
+    geom_point(data = filter(j, cum_rich > 2300), shape = 1) +
+    stat_smooth(aes(group = Mouse, lty = Cohoused), se = FALSE, data = filter(j, cum_rich < 2300)) +
+    scale_color_manual(values = c('red','black')) +
+    facet_wrap(~Abx) +
+    expand_limits(y = 0) +
+    th +
+    labs(x = "Hour", y = 'Cumulative no. of unique OTUs encountered')
+
+plot_it(p, name = 'figCumulativeOTUs', dir = image_dir, width = 6, height = 5)
+
+p
+
+```
+
+(ref:figCumulativeOTUs) Figure `r refs$figCumulativeOTUs`
+
+`r pagebreak` ![](images\\figCumulativeOTUs.png)
+
+**(ref:figCumulativeOTUs):**
+
 `r pagebreak` Works Cited

### 2018-09-06 custom_functions.R

@@ -3,24 +3,18 @@
 bac_extract <- function(x, trait) {
   
   # for each species query result, filter blanks and pick best value
-  for(i in 1:length(tax_bacdat)) {
-    names(x[[i]]) <- tax_bacdat[[i]]
-    x[[i]] <- x[[i]][sapply(x[[i]], length) > 0]
-    x[[i]] <- sapply(x[[i]], pick_best)
-  }
-  
-  # unlist, clean, return df
+  x <- lapply(x, unlist)
+  x <- lapply(x, pick_best)
   x <- unlist(x)
-  x[x == 'NULL'] <- NA
-  x_names <- strsplit(names(x), '.', fixed = TRUE)
+  x <- x[!is.na(x)]
   x <- data.frame(
-    Genus = sapply(x_names, function(y) y[[1]]),
-    Species = sapply(x_names, function(y) y[[2]]),
+    Genus = tax_bacdat$Genus[match(names(x), tax_bacdat$id)], 
+    Species = tax_bacdat$Species[match(names(x), tax_bacdat$id)],
     trait = trait,
-    val = x
+    val = x,
+    row.names = NULL,
+    stringsAsFactors = FALSE
   )
-  x <- x[!is.na(x$val), ]
-  x$val <- as.character(x$val)
   return(x)
 }
 

### 2018-09-06 mouse_analysis_sandbox.R

@@ -1,0 +1,19 @@
+###############################################################
+################# PLAYING WITH TREES #####################
+###############################################################
+
+#load taxonomies for labeling tree and exploring results.
+
+tax$source <- ifelse(paste(tax$Genus, tax$Species) %in% paste(tax_mouse$Genus, tax_mouse$Species), 'AR', 'LTP')
+iwant <- c(tax$otu[tax$source == 'AR'], sample(tax$otu[tax$source != 'AR'], 3*sum(tax$source == 'AR')))
+tmp <- drop.tip(tre, tre$tip.label[!tre$tip.label %in% iwant])
+
+#color by data source
+tmp <- groupOTU(tmp, grep('OTU', tmp$tip.label))
+
+#color by Genus
+ggtree(tmp, aes(color=group))
+ggtree(tmp, aes(color=group), layout='circular') +
+  theme(legend.position = c(0.5,0.4), 
+        legend.title = element_blank(), # no title
+        legend.key = element_blank()) # no keys

### 2018-09-06 mouse_fastas_to_tree.sh

@@ -1,0 +1,39 @@
+#Mouse project script: making a phylogenetic tree with samples
+
+# we will use living tree project -- very well annotated and small (which is a feature)
+wget "https://www.arb-silva.de/fileadmin/silva_databases/living_tree/LTP_release_132/LTPs132_datasets.fasta.tar.gz"
+tar -xvf LTPs132_datasets.fasta.tar.gz
+
+#standardize the format with output from the usearch pipeline
+#remove blank spaces
+sed '/^[^>]/ 's/[[:space:]]//g'' LTPs132_SSU_compressed.fasta > LTP.fasta
+
+#remove line breaks...
+sed -i ':a; $!N; /^>/!s/\n\([^>]\)/\1/; ta; P; D' LTP.fasta
+
+#now convert rna to dna (even though it doesn't matter for cluster -- i checked)
+sed -i '/^[^>]/ y/uU/tT/' LTP.fasta
+
+#now pick out just id from the header
+sed -i 's/\t.*//' LTP.fasta
+
+#now use cutadapt to select just the v4 region
+#i got these sequences from the GLBRC pipeline: https://github.com/GLBRC-TeamMicrobiome/TagAnalysis/blob/master/16S_USEARCH_Pipeline.md
+#note that we are using the first one in the 3' direction, and the second in the 5' direction
+module load cutadapt/1.8.1
+cutadapt -a ATTAGAWACCCBDGTAGTCC -o tmp.fasta LTP132.fasta > cut_Fadpt_results_LTP.txt
+cutadapt -g GTGCCAGCMGCCGCGGTAA -o LTP132.fasta tmp.fasta > cut_Radpt_results_LTP.txt
+
+#remove sequences that are less than 250, then trim sequences to 250
+awk 'BEGIN{RS=">";ORS=""}length($0)>249{print ">"$0}' LTP132.fasta > tmp.fasta
+cut -c 1-250 tmp.fasta > LTP132.fasta
+
+#combine LTP and mouse data
+cat LTP.fasta otus.fa > LTP_mouse.fa
+
+#calculate distance matrix (~30min)
+/mnt/research/rdp/public/thirdParty/usearch10.0.240_i86linux64 -calc_distmx LTP_mouse.fa -distmxout LTP_mouse_mx.txt
+
+#turn distance matrix into a tree (~1hr)
[…]

### 2018-09-06 mouse_microbiome_processing_trait_data_2.R

@@ -1,0 +1,713 @@
+#Mouse gut microbiome traits -- data gathering and cleaning
+#John Guittar
+
+##Overview
+#1. load raw taxonomic data from usearch output
+#2. load taxonomy from LTP (Living Tree Project)
+#3. remove any ambiguities/special characters
+#4. gather and combine trait data from various sources
+#5. filter trait data to only relevant taxa (either from usearch or LTP)
+#6. load tree of both usearch and LTP taxa, created with usearch -calc_distmx and -cluster_aggd
+#6.5 Do not infer unkonwn trait values yet! that's for another script
+#7. save clean taxonomic, trait, and phylogenetic data
+
+######################################################################
+######################################################################
+######################################################################
+
+#load taxonomic data from usearch output
+#remove any Genus/Species names with numbers or special characters
+#set working directory, load packages
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+source('C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\custom_functions.R')
+loadpax(c('data.table','tidyverse','castor','phytools'))
+
+#load raw taxonomy data from usearch
+#meta <- readRDS('metadata_aug19.RDS')
+tax <- read.table('combined_merged_otus_taxonomy.sintax', fill = TRUE) %>%
+  select(otu = V1, tax = V4) %>%
+  separate(tax, c('Kingdom','Phylum','Class','Order','Family','Genus','Species'), sep = ',', fill = 'right') %>%
+  mutate_all(funs(gsub('\\w:', '', .))) %>%
+  mutate_all(funs(ifelse(is.na(.), 'unclassified', .)))
+
+# Clean up taxonomy names -- no ambiguous genera names (important when querying bacdive)
+library(stringr)
+tax$Genus <- unlist(lapply(strsplit(tax$Genus, '_'), function(x) if(x[1] %in% tax$Family) x[2] else x[1]), tax$Genus)
[…]

### 2018-09-06 mouse_microbiome_processing_trait_data_3.R

@@ -1,0 +1,712 @@
+#Mouse gut microbiome traits -- data gathering and cleaning
+#John Guittar
+
+##Overview
+#1. load raw taxonomic data from usearch output
+#2. load taxonomy from LTP (Living Tree Project)
+#3. remove any ambiguities/special characters
+#4. gather and combine trait data from various sources
+#5. match trait data to as many taxa as possible, either from usearch or LTP
+#6. save raw trait data (no phylogenetic inference yet) 
+
+######################################################################
+######################################################################
+######################################################################
+
+#set working directory, load packages
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+source('C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\custom_functions.R')
+loadpax(c('data.table','tidyverse'))
+
+#load raw taxonomy data from usearch
+#Note: I also tried to use LTP to assign taxonomy. But it has WAY more 'unclassified'. Since we're just interested in getting taxon traits by mining the literature -- not on taxonomy/phylogeny per se -- we get way more hits if we use the whole SILVA database. Later, we map the OTUs onto the LTP phylogeny based on 16S similarity.
+tax_seqs <- read.table('combined_merged_otus_taxonomy.sintax', fill = TRUE) %>%
+  select(otu = V1, tax = V4) %>%
+  separate(tax, c('Kingdom','Phylum','Class','Order','Family','Genus','Species'), sep = ',', fill = 'right') %>%
+  mutate_all(funs(gsub('\\w:', '', .))) %>%
+  select(-Kingdom, -Phylum, -Class, -Order, -Family)
+
+#load raw taxonomy data from LTP version 132
+#originally, https://www.arb-silva.de/fileadmin/silva_databases/living_tree/LTP_release_132/LTPs132_SSU.csv
+#BUT: i used excel to quickly fix some of the formatting issues, and saved it as a csv
[…]

### 2018-09-06 mouse_processing_trait_data.R

@@ -1,0 +1,643 @@
+#Mouse gut microbiome traits -- data gathering and cleaning
+#John Guittar
+
+##Overview
+#1. load raw taxonomic data from usearch output
+#2. load taxonomy from LTP (Living Tree Project)
+#3. remove any ambiguities/special characters
+#4. gather and combine trait data from various sources
+#5. match trait data to as many taxa as possible, either from usearch or LTP
+#6. save raw trait data (no phylogenetic inference yet) 
+
+######################################################################
+######################################################################
+######################################################################
+
+#set working directory, load packages
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+source('C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\custom_functions.R')
+loadpax(c('data.table','tidyverse'))
+
+#load raw taxonomy data from usearch
+#Note: I also tried to use LTP to assign taxonomy. But it has WAY more 'unclassified'. Since we're just interested in getting taxon traits by mining the literature -- not on taxonomy/phylogeny per se -- we get way more hits if we use the whole SILVA database. Later, we map the OTUs onto the LTP phylogeny based on 16S similarity.
+tax_mouse <- read.table('combined_merged_otus_taxonomy.sintax', fill = TRUE) %>%
+  select(otu = V1, tax = V4) %>%
+  separate(tax, c('Kingdom','Phylum','Class','Order','Family','Genus','Species'), sep = ',', fill = 'right') %>%
+  mutate_all(funs(gsub('\\w:', '', .))) %>%
+  mutate(
+    Genus_original = Genus,
+    Species_original = Species,
+    Genus = trimws(Genus),
+    Genus = ifelse(Genus == 'Incertae_Sedis', NA, Genus),
+    Genus = gsub('\\[|\\]', '', Genus),
+    Genus = gsub('-.*', '', Genus),
+    Genus = gsub('Candidatus_', '', Genus),
[…]

### 2018-09-06 mouse_seqs_to_otus.sh

@@ -1,0 +1,96 @@
+#Mouse project script: going from raw seqs to an OTU table
+
+#make new directory and go there
+mkdir mouse
+cd mouse/
+
+#load most recent version of R
+module swap GNU GNU/4.9
+module load R/3.3.2
+R
+
+# download list of accessions
+download.file('https://www.ebi.ac.uk/ena/data/warehouse/filereport?accession=PRJEB26446&result=read_run&fields=study_accession,sample_accession,secondary_sample_accession,experiment_accession,run_accession,tax_id,scientific_name,instrument_model,library_layout,fastq_ftp,fastq_galaxy,submitted_ftp,submitted_galaxy,sra_ftp,sra_galaxy,cram_index_ftp,cram_index_galaxy&download=txt', 'Reese2018_accessions.txt')
+
+#read and simplify into list of URLS.
+urls <- read.table('Reese2018_accessions.txt', header = TRUE)
+urls <- unlist(strsplit(as.character(urls$submitted_ftp), ';'))
+urls <- paste0('http://', urls) 
+
+#save table, unquoted
+write.table(urls, 'Reese2018_accessions.txt', col.names = F, row.names = F, quote = F)
+
+#quit R
+quit(save = 'no')
+
+# download the files from ENA, uncompress
+wget -i Reese2018_accessions.txt
+gunzip ERR*.fastq.gz
+
+#save list of headers with sample names
+head -n 1 ERR*.fastq > headers.txt
+
+#R for data munging
+R
+
+#load list of sample headers (necessary to identify samples)
+heds <- readChar('headers.txt', file.info('headers.txt')$size)
+heds <- unlist(strsplit(heds, '==> '))
+heds <- heds[c(2:length(heds))]
+sample_names <- gsub(' .*', '', heds)
+ids <- gsub('.*\\.1 ', '', heds)
+ids <- gsub('.*\\.1 ', '', ids)
+ids <- gsub('_.*', '', ids)
+unwanted_fastqs <- paste(sample_names[grepl('D|swab', ids)], collapse = ' ')
+sample_dict <- data.frame(sample_name = sample_names, sample_id = ids)
+
+#write
+write(unwanted_fastqs, 'unwanted.fastqs.txt')
+write.csv(sample_dict, 'sample_dict.csv')
+
+#back to bash
+quit(save = 'no')
+
+#remove unwanted fastqs
+xargs rm < unwanted.fastqs.txt
+
[…]

### 2018-09-06 mouse_trait_predictions.R

@@ -1,0 +1,337 @@
+#Mouse gut microbiome traits -- inferring unknown traits using phylogeny
+#John Guittar
+
+#load packages and set working directory
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+source('C:\\Users\\John\\Documents\\msu\\Mouse_Microbiome\\custom_functions.R')
+loadpax(c('tidyverse','castor','phytools','stringr','ggtree','phylosignal','kableExtra'))
+
+#load tree with LTP and mouse tips, calculated with usearch
+tre <- read_tree(write.tree(read.tree('LTP_mouse.tree')))
+
+#load traits
+traits <- as.data.frame(readRDS('traits_sparse.RDS')) %>%
+  mutate(binomial = paste(Genus, Species))
+
+trait_names <- c(
+  "Aggregation_score" = 'Aggregation score',
+  "B_vitamins" = "B vitamins",       
+  "Copies_16S" = "16S gene copies",
+  "GC_content" = "GC content",       
+  "Gene_number" = "Genes",
+  "Genome_Mb" = "Genome size",
+  "Gram_positive" = "Gram-positive",     
+  "IgA" = 'IgA binding affinity',             
+  "Length" = "Length",
+  "Motility" = "Motility",
+  "Oxygen_tolerance" = "Oxygen tolerance",
+  "pH_optimum" = "pH optimum",
+  "Salt_optimum" = "Salinity optimum",      
+  "Sporulation" = "Sporulation score",      
+  "Temp_optimum" = "Temperature optimum",
+  "Width" = "Width"
+)
+
+#load taxonomy file (used to link otus and traits)
+#files were created in mouse_processing_trait_data.R
+tax_LTP <- readRDS('LTP_tax.RDS')
+tax_mouse <- readRDS('mouse_tax_SILVA.RDS') %>%
+  select(otu, Genus, Species)
+tax <- bind_rows(tax_LTP, tax_mouse) %>%
+  mutate(binomial = paste(Genus, Species))
+
+#load OTU table for plotting abundance-weighted coverage
+otus <- readRDS('otu_table_usearch_aug19.RDS')
+
+#join traits and OTUs using curated latin binomials
+tax_traits <- tax %>% left_join(select(traits, -Genus, -Species), by = c("binomial"))
+
+##Estimating unknown trait values, and evaluating confidence in those estimations
[…]

### 2018-09-06 processesing_mouse_metadata.R

@@ -1,0 +1,90 @@
+wd <- 'C:\\Users\\John\\Google Drive\\Mouse_Microbiome_Shared_Files\\Data\\'
+setwd(wd)
+
+#load otu data
+x <- fread('mouse_combined_merged_OTU_table.txt') %>%
+  rename(otu = `#OTU ID`) %>%
+  gather(sample, abun, -otu)
+
+#load metadata 
+x1 <- fread('Reese_2018_04_29_20151231_meta.txt') %>% as.data.frame()
+names(x1) <- x1[3,]
+x1 <- x1[c(6:nrow(x1)), ]
+
+x2 <- fread('Reese_2018_04_29_20150930_meta.txt') %>% as.data.frame()
+names(x2) <- x2[3,]
+x2 <- filter(x2, sample_alias != '')
+x2 <- x2[c(6:nrow(x2)), ]
+
+#merge metadata
+
+cutsub <- function(strings, pattern, cut) {
+  # function to extract a substring in a vector of strings...
+  tmp <- str_extract(strings, pattern)
+  # then trim that extraction...
+  tmp <- gsub(cut, '', tmp)
+  #then convert to numeric...
+  tmp <- as.numeric(tmp)
+  return(tmp)
+}
+
+meta <- rbind(x1, x2) %>%
+  rename(x = sample_description) %>%
+  transmute(sample = sample_alias, 
+            mouse = cutsub(x, "mouse \\d+", "mouse "),
+            cohoused = grepl('cohoused=y', x),
+            partner = cutsub(x, "cohoused=y with mouse \\d+", 'cohoused=y with mouse '),
+            pair = as.numeric(factor(pmax(mouse, partner))),
+            hour = ifelse(grepl('day', x), cutsub(x, "day \\d+", 'day ') * 24,
+                     ifelse(grepl('hour', x), cutsub(x, "hour \\d+", 'hour '), 0)),
+            cells_per_gram = as.numeric(`organism count`) / as.numeric(`amount or size of sample collected`), 
+            Abx = `miscellaneous parameter` == 'antibiotic')
+
+
+#remove samples that are in metadata but did not sequence correctly
+meta <- filter(meta, !sample %in% as.character(c(1.02, 1.097, 10.087, 11.087, 13.087, 14.087, 15.073, 16.073, 3.02, 5.087, 8.008)))
+
+#fix another mapping error
+meta$hour[meta$mouse < 20 & meta$hour == 144] <- 120
[…]
